---
title: "Stat 33B - Lecture 6"
date: February 26, 2020
output: pdf_document
---

Announcements
=============

Quiz 2 this Friday:

* Similar length to Quiz 1
* Multiple-choice and fill-in-the-blank questions.
* Covers everything up to the end of last week.


Review
======

Previous lecture:

* Tidy data
* The `pivot_wider` function


We took this untidy data:
```{r}
library(tidyr)

table2
```

And used the `pivot_wider()` function to rotate rows into columns:

```{r}
pivot_wider(table2, names_from = type, values_from = count)
```


Rotating Columns into Rows
--------------------------

Some data sets put measurements for multiple observations in a single
row.

In this table, each row contains the measurements for two
observations of the `cases` variable:
```{r}
table4a
```

To make this data tidy:

* Rotate the `1999` and `2000` columns into rows.
* New columns are `year` and `cases`.

Tidying this data set makes it longer.


__Important Note:__ Earlier we saw negative indexing with positions:
```{r}
table4a[-1]
```

Tidyverse functions (but not base R!) also support negative indexing with
names!


Use `pivot_longer()` to rotate columns into rows:
```{r}
pivot_longer(table4a, -country, names_to = "year", values_to = "cases")
```

Need to specify:

* Columns to rotate as `cols`.
* Name(s) of new identifier column(s) as `names_to`.
* Name(s) of new measuerment column(s) as `values_to`.


See `vignette("pivot")` for more examples of using `tidyr`.


If you wanted to do this without `tidyr`:

1. Subset columns to separate `1999` and `2000` into two data frames.
2. Add a `year` column to each.
3. Rename the `1999` and `2000` columns to `cases`.
4. Stack the two data frames with `rbind()`.

```{r}

col99 = table4a[-3]
col00 = table4a[-2]

names(col99)[2] = "cases"
col99["year"] = 1999

col99

names(col00)[2] = "cases"
col00["year"] = 2000

col00

rbind(col99, col00)
```


Merging Data
============

A merge or "join" combines data from two separate data frames, based
on some identifying values they have in common.


Recall the `pivot_wider()` example with `table2`.

To make `table2` tidy without `tidyr`:

1. Subset rows to separate `cases` and `population` values.
2. Remove the `type` column from each.
3. Rename the `count` column to `cases` and `population`.
4. Merge these two subsets by matching `country` and `year`.

The two subsets are:
```{r}
table2

cases = table2[table2$type == "cases", ]
pop = table2[table2$type == "population", ]

names(pop)[4] = "population"
pop = pop[-3]

pop

names(cases)[4] = "cases"
cases = cases[-3]
```


BAD ways to combine rows:

```{r, eval = FALSE}
cases$population = pop$population

# OR

cbind(cases, pop["population"])
```

Dangerous if you're not 110% sure the order of rows matches!


Imagine if the rows were in a different order:
```{r}
pop = pop[c(4, 5, 1, 2, 3, 6), ]
```

Then adding a new column or using `cbind()` would mix up the observations.


Instead, match rows on `country` and `year` with the `merge()` function:
```{r}
merge(cases, pop)
```

The function automatically uses columns that have the same name in both tables
to match.


More Merges
-----------

Data split across multiple tables are called _relational data_.

A column shared by several tables is called a _key_.

For example, a grocery store's inventory system might have:

* A table that lists stores
* A table that lists items (fruits, vegetables, etc)
* A table that lists quantity of each item at each store

See the bCourse for the data set:
```{r}
stores = readRDS("data/grocery/stores.rds")

stores

items = readRDS("data/grocery/items.rds")

items

inv = readRDS("data/grocery/inventory.rds")

inv
```

By default, the `merge()` function only keeps rows that match:

```{r}
merge(stores, inv)
```

Use the `all` parameter to force all rows from both tables to show
up:

```{r}
merge(stores, inv, by.x = "store_id", by.y = "StoreID", all = TRUE)
```

There are also `all.x` and `all.y` parameters to only force all rows
from one table.


Use the `by` parameter to specify the key. If the name of the key is
different for the two tables, use `by.x` and `by.y`.

```{r}
merge(stores, inv, by.x = "store_id", by.y = "StoreID")
```

If you're familiar with SQL `JOIN`s, the `merge()` function is the
same idea. Specifically:

* `INNER JOIN` is `all = FALSE` (the default)
* `LEFT JOIN` is `all.x = TRUE`
* `RIGHT JOIN` is `all.y = TRUE`
* `OUTER JOIN` is `all = TRUE`
