---
title: "Stat 33B - Lecture 7"
date: March 4, 2020
output: pdf_document
---

Announcements
=============

Due to the coronavirus, I'm considering alternatives to in-person
lectures/labs and the attendance grade.

Homework 3 due today.

Homework 4 will be posted tomorrow.


Review
======

Grocery data set on bCourse:
```{r}
stores = readRDS("data/grocery/stores.rds")
items = readRDS("data/grocery/items.rds")
inv = readRDS("data/grocery/inventory.rds")

stores

inv
```

A merge or "join" combines data from two separate data frames, based
on some identifying values they have in common:
```{r}
merge(items, inv)
```


If-statements
=============

If-statements run some code only if a condition is true.

Three ways to write if-statements in R.

1. Ordinary if-statements

```{r}
x = 7

if (x > 5)
  message("hello!")
```

Use `{`, the curly brace, to write multiple line if-statements.
```{r}
if (x > 5) {
  message("hello!")
  message("x is greater than 5!")
}
```

Use `else` to add code that only runs if the condition is false:
```{r}
if (x < 5) message("x is less than 5!") else message("x is greater than 5!")
```

Note that if you spread an if-else statement across multiple lines, you need to
include `else` on the last line of the if-statement:
```{r}
if (x < 5)
  message("x is less than 5!") else
  message("x is greater than 5!")

# Using curly braces:
if (x < 5) {
  message("x is less than 5!")
} else {
  message("x is greater than 5!")
}
```
R ignores whitespace, but it is idiomatic to indent code inside of an
if-statement.


Multiple if-statements can be chained together:
```{r}
if (x < 5) {
  message("hello!")
  message("x is less than 5!")
} else if (x %% 2 == 0) { # if x is even
  message("x is even!")
} else {
  message("x is greater than or equal to 5, and odd!")
}
```


If-statements return the value of their last expression:
```{r}
x = 28
y = if (x < 5) 42 else 8

y = if (x < 5) 42 # else NULL

y = if (x < 5) {
  47
  32
  31
}

y = {
  61
  32
  51
}
```
When there is no `else` block, the value of the `else` block is `NULL`.


2. The `ifelse()` function is a vectorized if-statement.
  
```{r}
x = c(1, 2, 3)
if (x < 2) {
  message("x is less than 2!")
}

ifelse(x < 2, x, 10)
```

Avoid `ifelse()` unless you need vectorization. Slower than ordinary
if-statement.


3. The `switch()` function

Matches an integer or string to one of several possible branches.

```{r}
# With a string:
x = "33b"

switch(x,
       "hello" = message("You said hello!"),
       "bye" = {
         message("Leaving so soon?")
         message("I hoped you'd stay")
       },
       # Default case
       message("Your variable didn't match")
)


# With an integer, matches by position:
x = 1

switch(x,
       message("x is 1"),
       message("x is 2"))

```

If you're using `switch()` to return values, you could use a list instead:
```{r}
switch(x, 20, 25)

mycases = list(20, 25)
mycases[[x]]

# Only 20 is evaluated:
#
#   switch(x, 20, time_consuming_function())
# 
# Both 20 and time_consuming_function() are evaluated:
#
#   list(20, time_consuming_function())
#
```
However, `switch()` is generally faster, because it only evaluates the code for
the branch that matches


Loops
=====

Loops run some code repeatedly.

Four ways to write loops in R:

1. Vectorization (fast!)
2. Apply functions: `lapply()`, `sapply()`, etc
3. `for`, `while`, and `repeat` loops
4. Recursion (slow)

Vectorization
-------------

A "vectorized" function is one that automatically applies an
operation to each element of a vector.

We saw this earlier with arithmetic:
```{r}
sin(c(1, 2, 3))
```

Apply Functions
---------------

Similar to "map" functions in other languages.

Applies a function (FUN) to each element of a vector or list (X):
```
lapply(X, FUN, ...)
```


`lapply()` always returns the result as a list:
```{r}
mydata = list(c(1:3, NA), seq(10, 20, 0.5), rnorm(10))

mean(mydata[[1]], na.rm = TRUE)
mean(mydata[[2]])

lapply(mydata, mean, na.rm = TRUE)

lapply(c(1, 2, 3), sin) # vectorization would be better

dogs = readRDS("data/dogs/dogs_full.rds")
# lapply() keeps the names of the input
lapply(dogs, class)
```
"l" for **list**.


Most apply functions will accept additional arguments to the applied
function. These arguments are held constant across all elements of
the data:
```{r}

```


`sapply()` simplifies the result to a vector, when possible:
```{r}
sapply(dogs, class)

sapply(mydata, mean, na.rm = c(TRUE, FALSE, FALSE))
```
"s" for **simplified** result.


`vapply()` simplifies the result to a vector of a specific type:
```{r}
vapply(dogs, class, "")
```
"v" is for **vector** result.



`mapply()` applies a function to each element of multiple vectors or
lists:
```{r}
x = 1:3
y = 2:4

x + y

# Same thing, less efficiently:
mapply(`+`, x, y)

# A better example:
mydata1 = list(1:3, 4:6)
mydata2 = list(4, 5:10)

mapply(sum, mydata1, mydata2)

# Another example:
missing = list(1:2, c(1:3, NA), c(1:4, NA))
na.rm = c(TRUE, FALSE, TRUE)
mapply(mean, missing, na.rm = na.rm)
```
"m" for **multiple** vectors.

---

_This is where the lecture ended. I've included the rest of the notes on loops
for anyone that wants to read ahead. The examples here will not necessarily be
identical to the ones I use in the next lecture._

### Flashback to Data Analysis

The `split()` function splits a vector or data frame into groups
based on some other vector.

Split weight of dogs by the group column:
```{r}
split(dogs$weight, dogs$group)
```

The `split()` function is especially useful when combined with
`lapply()` or `sapply`().

```{r}
by_group = split(dogs$weight, dogs$group)
sapply(by_group, mean, na.rm = TRUE)
```
This is an R idiom!


The `tapply()` function is equivalent to the `split()` and `sapply()`
idiom:
```{r}
tapply(dogs$weight, dogs$group, mean, na.rm = TRUE)
```
"t" for **table**, because `tapply()` is a generalization of the
frequency-counting function `table()`.


The `split()` and `tapply()` functions are important for analyzing
data in data frames.


### Other Apply Functions

* `apply()` applies a function to rows, columns, or elements of a
  matrix, and returns a matrix.
* `eapply()` applies a function to the elements of an environment.
  We'll learn about environments in a later lecture.
* `rapply()` applies a function to all elements of a recursive list
  (a list of lists).

See this StackOverflow Post for a good summary:

    https://stackoverflow.com/a/7141669



`for`, `while`, and `repeat`
----------------------

A for-loop runs a block of code once for each element of a vector or
list:
```{r}
for (i in 1:10) {
  message(i)
}
```
The same idea as for-loops in other languages.


A while-loop runs a block of code repeatedly as long as some
condition is true:
```{r}
i = 1
while (i < 10) {
  message(i)
  i = i + 1
}
```
Again, the same idea as in other languages.


Use `break` to exit a loop early:
```{r}
for (i in 1:10) {
  message(i)
  if (i == 3)
    break
}
```

Use `next` to skip to the next iteration early:
```{r}
for (i in 1:10) {
  if (i %% 3 == 0) # skip numbers divisible by 3
    next

  message(i)
}
```


A repeat-loop runs a block of code repeatedly:
```{r}
i = 1
repeat {
  message(i)

  if (i == 10)
    break

  i = i + 1
}
```

If you know how many results you expect to get from the loop,
allocate storage before the loop begins:
```{r}
result = numeric(3)
a_vec = c(1, 1, 1)
b_vec = c(0, 4, 2)
c_vec = c(-16, -12, -8)

for (i in 1:3) {
  a = a_vec[i]
  b = b_vec[i]
  c = c_vec[i]

  result[i] = (-b + sqrt(b^2 - 4 * a * c)) / (2 * a)
}
```

If you need indices, using `1:length(x)` can cause bugs:
```{r}
x = c()

1:length(x)
```

Use `seq_along(x)` instead:
```{r}
seq_along(x)
```


Recursion
---------

_See the lecture notes for next week._


How to Write Loops
------------------

When thinking about writing a loop, try (in order):

1. vectorization
2. apply functions
    * Try an apply function if iterations are independent.
3. for/while-loops
    * Try a for-loop if some iterations depend on others.
    * Try a while-loop if the number of iterations is unknown.
4. recursion
    * Convenient for naturally recursive problems (like Fibonacci),
      but often there are faster solutions.

Before you write the loop, try writing the code for just 1 iteration.

Make sure that code works; it's easy to test code for 1 iteration.

When you have 1 iteration working, then try using the code in a loop
(you will have to make some small changes).

If your loop doesn't work, try to figure out which iteration is
causing the problem. One way to do this is to use message() to print
out information.

Then try to write the code for the broken iteration, get that
iteration working, and repeat this whole process.
