---
title: "Stat 33A - Lecture 3"
date: February 24, 2020
output: pdf_document
---

Announcements
=============

Homework 2 will be posted later today.


Review
======

## Implicit Coercion

Last time we saw that R can automatically convert between ("coerce")
types in one direction:

    logical -> integer -> numeric -> complex -> character


```{r}
class(5.1)

x = "hello"
class(x)

TRUE + 5

c(6, 8, 1)

c("hi", "hello")

c(7, "hi")
```

## Lists

In a vector, every element has to have the same type.

A list is a container for elements with *different* types.

If you use elements that can't be coerced to a common type, `c()`
creates a list instead:

```{r}
class(sin)

class(c(5, sin))
```

You can also directly create a list with the `list()` function:
```{r}
list(5, "hi", 6.1)

x = list(5, 6)

class(x)

# Lists print with [[ to indicate element positions.
x 


# List elements can have names. Named elements print with $ instead:
list(a = 1, 2)
```

Vectorized operations don't work for lists:
```{r, error = TRUE}
c(1, 2) + c(3, 4) # ok

list(1, 2) + list(3, 4) # not ok
```


Projects and Files
==================

## Setting Up A Data Analysis

1. Create a project directory.
2. Download the data to the project directory.
3. Read the data into R.

For this class, think of each assignment as a new project.


## The File System Tree

In order to read a data set, you need to tell R where it is on your
computer.

Your computer's file system is like a tree.

The root is at `/` (OS X, Linux) or `C:/` (Windows).

Each directory is a branch.


## Paths

The "path" to a file is the sequence of directories it's in,
separated by forward slashes `/`.

For example, the file `dinosaur.csv`, in the directory `data,` in the
directory `storage`, in the root directory:
```
/storage/data/dinosaur.csv
```

Windows traditionally uses backslashes instead. You can still use
forward slashes in R.


An "absolute path" is one that starts from the root directory.


## Paths in R

Absolute paths can be infuriatingly long to type.

You can set a "working directory" in R and then use "relative paths"
that start from the working directory.

Use `getwd()` to check the working directory:
```{r}
getwd()
# Check the PDF version of the notes if you want to see the output on
# my computer.
```

The output on your computer will be different, because you have
different files!


__PITFALL__: RStudio maintains the working directory for your console
independently of the working directory for each Rmd notebook. So:

* Running `getwd()` from the notebook with `Ctrl + Enter` displays
  the NOTEBOOK'S working directory.
* Typing `getwd()` in the "Console" window and pressing `Enter`
  displays the CONSOLE's working directory.

In the console, you can use `setwd()` with a path to set the working
directory.

In the notebook, if you use `setwd()` it only lasts for __that
chunk__ and is then reset:
```{r}
setwd("/home/nick/university/teach/stat33ab/stat33b")

getwd()
```

So in subsequent chunks it looks like you didn't call `setwd()`:
```{r}
getwd()
```

Why does RStudio do this? It is a bad practice to include `setwd()`
in your notebooks, because people you share the notebook with, like
your colleagues, instructor, or employer, might not have the same
directories on their computer as the ones you have on your computer.
The next section has more details about this.

By default, RStudio does the right thing and sets the notebook's
working directory to the place where the notebook is saved. Then you
can use relative paths (see below) to load and save files from the
notebook.

If you really want to set the working directory in a notebook, it is
possible to override RStudio. See <https://yihui.org/knitr/options/>
for details.

Use `list.files()` to list files in a directory:
```{r}
list.files()
```

A "relative path" is one that starts from the working directory.
```{r}
# List the files in the "data" directory:
list.files("data")

# The relative path "data" stands for the absolute path
#
# "/home/nick/university/teach/stat33ab/stat33a/sandbox/data"
#
# So it saves a lot of typing!
```

If you see `character(0)` as the output from `list.files()`, that
means one of:

* The path you provided is incorrect.
* The path you provided leads to a file, not a directory.
* There are no files in the directory.

For example, if we make a deliberate typo:
```{r}
list.files("dat") # no files, the path is incorrect
```

The path `..` is a shortcut for the directory above:
```{r}
getwd()

list.files("..") # what's in the directory above

# You can use .. more than once:
list.files("../..") # what's in the directory two levels above

# You can also use .. with regular directory names:
list.files("../../stat33b") # two levels above, then back down
```


The path `~` means your personal directory:
```{r}
# `~` is `/home/nick` on my computer.
#
# It will be different on your computer.

list.files("~") # lists files in /home/nick
```

You can convert a relative path to an absolute path with
`normalizePath()`:
```{r}
# Where is `~`?
normalizePath("~")
```


## Reproducible Analyses

Plan ahead so that other people can run your code and reproduce your
results.

Good habits:

* Putting your notebook(s) and data in the project directory.
* Using paths relative to the project directory.

Bad habits:

* Calling `setwd()` in R notebooks and scripts.
* Using absolute paths.

It's okay to use `setwd()` in the *R console* to set the working
directory to your project directory.



Working with Tabular Data
=========================

## Data Frames

In statistics, tabular data usually has:
  * Observations as rows
  * Features as columns

R's data structure for tabular data is the "data frame".

For the next few lectures, we'll use the Dogs Data Set, from:

    https://informationisbeautiful.net/visualizations/
        best-in-show-whats-the-top-data-dog/


Also posted on the bCourse in RDS format.

You can read an RDS file with `readRDS`:
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
```

Usually not a good idea to print an unfamiliar data set:
```{r, eval = FALSE}
# In the Rmd version of the notes, I've set eval = FALSE here so that
# this chunk doesn't actually run or print anything.

dogs # don't do this with unfamiliar data sets
```
Printing a large data set will also slow down knitting for R
notebooks!

Instead, inspect the data set with functions.

We already saw one function for inspecting data:
```{r}
class(dogs) # note it is a data.frame
```


Use `head()` to print the first 6 rows (or elements):
```{r}
head(dogs, 4)
```

Use `tail()` for the last 6:
```{r}
tail(dogs)
```

Use `dim()` to print the dimensions:
```{r}
dim(dogs)
```

Alternatively, use `ncol()` and `nrow()`:
```{r}
ncol(dogs)

nrow(dogs)
```

Use `names()` to print the column (or element) names:
```{r}
names(dogs)
```

Use `rownames()` to print the row names:
```{r}
rownames(dogs)
```

Use `str()` to print a structural summary:
```{r}
str(dogs)
```

Use `summary()` to print a statistical summary:
```{r}
summary(dogs)
```

In a data frame:

* Every column must be the same length.
* Every row must be the same length.
* Each column must have homogeneous elements (like a vector).
* Each row may have heterogeneous elements (like a list).

---

_This is where the lecture ended.

I've included some additional notes to help you get started on the
next homework assignment. These will also be covered in the lab and
in the next lecture._

## R Packages

The Comprehensive R Archive Network (CRAN) is a repository of
user-contributed packages for R.

You can install packages from CRAN with `install.packages()`:
```{r, eval = FALSE}
install.packages("dplyr")
```

For maintaining your packages, there are also functions:

* `remove.packages()` to remove a package
* `update.packages()` to update ALL packages
* `installed.packages()` to list installed packages

You can load an installed package into your R session with the
`library()` function:
```{r}
library(dplyr)
```

The process:

* Use `install.packages()` to install a package the **first time**
  you want to use the package, or if you want to update just one
  package to the latest version.

* Each time you start R, use `library()` to load the packages you
  want to use. This includes right after installing a package.


## dplyr

The `dplyr` package provides functions for working with data frames.

We'll use `dplyr` for now, and learn about R's built-in tools later.

Cheat sheet:

    https://github.com/rstudio/cheatsheets/raw/master/
        data-transformation.pdf


Use `slice()` to choose rows by position:
```{r}
slice(dogs, 1) # row 1

slice(dogs, 2) # row 2
```

Use `filter()` to choose rows that satisfy a condition:
```{r}
# Dogs with weight greater than 60:
filter(dogs, weight > 60)

# Dogs with weight greater than mean weight:
filter(dogs, weight > mean(weight, na.rm = TRUE))

# Dogs with breed equal to Bulldog:
filter(dogs, breed == "Bulldog")
```

Use `select()` to choose columns by name or position:
```{r}
# Get datadog column:
select(dogs, datadog)
```

Use `:` to indicate a range of rows or columns:
```{r}
# Get first 3 rows:
slice(dogs, 1:3)

# Get columns 2 through 4:
select(dogs, 2:4)

# Get columns breed through popularity:
select(dogs, breed:popularity)
```

Use `-` to exclude rows or columns:
```{r}
# Get all columns except breed:
select(dogs, -breed)

# Get all rows except 5 through 10:
select(dogs, -(5:10))

# Note that -5:10 is different from -(5:10) because of order of
# operations!
```


Other useful `dplyr` functions:

* `arrange()` changes the ordering of the rows.
* `mutate()` adds new columns by transforming existing columns.
* `summarise()` reduces multiple rows down to a single value.
* `group_by()` splits rows into groups when summarizing.
