---
title: "Stat 33B - Lecture 9"
date: March 18, 2020
output: pdf_document
---


Scope & Dynamic Lookup
======================

Scope
-----

The "scope" of a variable is the section of a program where that
variable exists and can be accessed.


In R:

* Function definitions are the primary way to create a new scope.
* If-statements and loops do not affect scope at all.


You can test whether a variable is in scope with the `exists()`
function:
```{r}
exists("x")

x = 3
x

exists("x")
```


Variables defined inside of a function are "local" to that function.

Local variables are not visible outside the function:
```{r}
f = function() {
  y = 44
  exists("y")
}

g = function() {
  y
}

f()
g()

y
```

Local variables reset each time the function is called:
```{r}
create_or_add = function() {
  if (exists("counter")) {
    counter = counter + 1
  } else {
    counter = 0
  }
  counter
}

create_or_add()
create_or_add()
```


A function can use variables defined outside (non-local), but only if
those variables are in scope where the function was **defined**.

This property is called "lexical scoping".

For example:
```{r}
getx = function() x

x

getx()

make_gety = function() {
  y = 3
  
  function() x + y
}

gety = make_gety()
gety()

test_gety = function() {
  y = 15
  gety()
}

test_gety()
```

Variables defined directly in the R console are "global" and
available to any function.


Local variables "mask" or hide non-local variables with the same name:
```{r}
getx_local = function() {
  x = "Hello"
  x
}

getx_local()
```
Locals get priority!


EXCEPTION: R ignores non-functions when looking up the name of a
called function.

For example:
```{r}
compute_mean = function() {
  x = c(1, 2, 3)
  mean = 0 
  # the local variable mean is ignored, because it's not a function
  mean(x)
}

compute_mean()
```


Besides function definitions, the `local()` function also creates a
new scope:
```{r}
local({
  z = 4
})
```


In summary:

* Function definitions (or `local()`) create a new scope.

* Local variables get reset each time a function is called.

* Where a function is **defined** determines which variables are in
  scope.

* Local variables mask non-local variables.

* R ignores non-functions when looking up the name of a called
  function.

For a function, **where** a variable will be looked for depends only
on where the function was defined (because of lexical scoping).


Dynamic Lookup, Part 1: Functions
---------------------------------

Variables are only looked up **when** they are actually used.

For a function, this means variables are only looked up **when** the
function is called.

This is called "dynamic lookup".

For example:
```{r}
get_cats = function() cats

cats = 3
get_cats()

cats = 21
get_cats()
```

Dynamic lookup can be counterintuitive.



Environments
============

The data structure that R uses to keep track of variables at run-time
is called an "environment".

Each environment has a "frame" that maps names to R objects (a hash
table).

Each environment also has a "parent environment" (with one exception
we'll see later).


Dynamic Lookup, Part 2: Environments
------------------------------------

At run-time, **each call** to a function creates a new environment:

* Its frame contains the function's local variables.

* Its parent environment is the environment where the function was
  **defined**. This satisfies lexical scoping.

When R looks up a variable, it checks the current environment first.
If the variable isn't there, it checks the environment's parent, then
the environment's parent's parent, and so on.


The "global environment" corresponds to the R console.

Use `globalenv()` to get the global environment:
```{r}
globalenv()
```


Assignment in Environments
--------------------------

The `<-` and `=` assign a variable in the current environment.

Use `<<-` to assign a variable in the parent environment:
```{r}
assignx = function(newx) {
  x <<- newx
}

x
assignx("Reassigned")
x
```
This means you can write a function that has a "side effect" on the
environment where it was defined.


In the R community, side effects are generally frowned upon, because
they make code harder to understand and predict.

Some side effects are useful:

* Making plots.
* Writing data to a file.

But side effects are not necessary for most functions, and you should
avoid them when possible.


Use `assign()` to assign a variable in a specific environment:
```{r}
assign("stat33", "is great", globalenv())
# stat33 = "is great"
stat33

assign("dogs", 3)
dogs
```

Most of R's functions for working with environments assume the
current environment if you don't specify an environment.


Unlike other R objects, environments **do not** follow the
copy-on-write rule.

They are reference objects:
```{r}
# Copy-on-write example:
x = c(4, 5, 6)
y = x
x[1] = 6
x
y

# Exception is environments:
genv = globalenv()
genv2 = genv
genv$x = "This is a test!"
genv$x
genv2$x
```


Inspecting Environments
-----------------------

Use `ls()` to list the names (variables) in an environment:
```{r}
ls()
```

By default, the `ls()` function ignores names that start with `.`

Use `all.names = TRUE` to make the function print these names:
```{r}
.x = 14

ls(all.names = TRUE)

ls(name = globalenv())
```

The `names()` function can also print out all names in an environment:
```{r}
names(globalenv())
```



Use `exists()` to check whether a variable is in an environment:
```{r}
exists("hi", globalenv())
```

Use `[[`, or `$`, `get()` to get a variable:
```{r}
genv = globalenv()

genv$cats
genv[["cats"]]

get("cats", globalenv())
```

Use `parent.env()` to get the parent environment:
```{r}
parent.env(globalenv())
```

Dynamic Lookup, Part 3: The Search Path
---------------------------------------

The global environment is not the top-level environment.

Besides your own code, R also uses environments to keep track of
packages.

When you load a package with `library()`, R creates a new environment:

  * Its frame contains the package's local variables.

  * Its parent environment is the envionment of the most recently
    loaded package.

  * The new environment becomes the parent of the global environment.

So R remembers the order in which packages are loaded.

This history of packages is called the "search path".

Use `search()` to see the names of environments in the search path:
```{r}
search()
```

The "base environment", or `package:base`, is a special environment
that R creates at startup.

The parent of the base environment is another special environment
called the "empty environment", which contains no variables and has
no parent.
```{r}
parent.env(baseenv())

parent.env(parent.env(globalenv())) # and so on..
```



Use `PACKAGE::NAME` to access a name in a specific package:
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
dplyr::filter(dogs, weight < 20)

# 1. Use to disambiguate which function you're calling.
# stats::filter()

# 2. To use just one function from a package.
dplyr::filter()
```
You do not have to load the package first to do this (but the package
does have to be installed).



Closures
--------

A "closure" is a function that keeps track of its environment.

The idea of a closure is used in many languages, not just R.

In R, every function is a closure.

Functions keep track of the environment where they were defined.

Use `environment()` to get the environment where a function was
defined:
```{r}
f = function() 42

environment(f)
```

You can use the closure property to create functions that remember
previous calls.

For example, suppose we want to make a counter function that keeps
track of how many times it's been called:
```{r}
make_counter = function() {
  count = 0
  
  function() {
    count <<- count + 1
    count
  }
}

counter = make_counter()
environment(counter)$count

counter()
counter()
```

Use functions with "memory" sparingly. They make it harder for others
to understand your code.
