---
title: "Stat 33A - Lecture 5"
date: March 9, 2020
output: pdf_document
---


Announcements
=============


Review
======

Last time:

The fundamental idea of `ggplot2` is that all graphics are composed
of layers.


Documentation:

    https://ggplot2.tidyverse.org/

Cheat sheet:

    https://github.com/rstudio/cheatsheets/raw/master/
        data-visualization-2.1.pdf


The Best in Show visualization:

    https://informationisbeautiful.net/visualizations/
        best-in-show-whats-the-top-data-dog/


```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
```

Layer 1: Data -- the data frame to plot.

Call the `ggplot()` function to set the data layer:
```{r}
library(ggplot2)

ggplot(dogs)
```

Layer 2: GEOMetry -- shapes to represent data.

Add a `geom_` function to set the geometry layer:
```{r}
ggplot(dogs) + geom_point()
```

Layer 3: AESthetic -- "wires" between geometry and data

Call the `aes()` function in `ggplot()` or the `geom_` to set the
mapping between the data layer and geometry layer:
```{r}
ggplot(dogs, aes(datadog, popularity_all)) + geom_point()

ggplot(dogs) + geom_point(aes(datadog, popularity_all))
```

Important layers in `ggplot2`:

Layer       | Description
----------  | -----------
data        | A data frame to visualize
aesthetics  | The map or "wires" between data and geometry
geometry    | Geometry to represent the data visually
statistics  | An alternative to geometry
scales      | How numbers in data are converted to numbers on screen
labels      | Titles and axis labels
guides      | Legend settings
annotations | Additional geoms that are not mapped to data
coordinates | Coordinate systems (Cartesian, logarithmic, polar)
facets      | Side-by-side panels


Saving Your Plots
-----------------

Use `ggsave()` to save the last plot you created to a file:

```{r}
ggsave("myplot.jpeg")

# The plot is saved in the notebook's working directory.
# You can check the working directory with:
getwd()
```

Building Up ggplot2 Plots
=========================

How else can we make our plot look more like the Best in Show plot?


1. Add the dog breeds as text.

Add more geometries to add additional details to a plot:
```{r}
ggplot(dogs, aes(datadog, popularity_all, label = breed)) +
    geom_point() +
    geom_text(size = 3, vjust = 0, hjust = 1, nudge_x = -0.05)
```

See the `ggrepel` package for automatic label positioning.


2. Color the points by type of dog.

```{r}
ggplot(dogs, aes(datadog, popularity_all, label = breed)) +
    geom_point(aes(color = group)) +
    geom_text(size = 3, vjust = 0, hjust = 1, nudge_x = -0.05)
```

__NOTE:__ Where you set the aesthetics matters. In the code above,
the `x` and `y` aesthetics apply to ALL geometries. The `color`
aesthetic only applies to `geom_point`, and the `label` aesthetic
only applies to `geom_text`. To make these aesthetics apply to
everything, move them to the `aes` call in the data layer.


We can also set parameters outside of the aesthetics.

Doing so sets a constant value instead of mapping to a feature in the
data.

Set size to 10 for all points:
```{r}
ggplot(dogs, aes(datadog, popularity_all)) +
    geom_point(aes(color = group), size = 10)
```

Note that if you want to set a constant color for all points, you need to do so
outside of `aes()`:
```{r}
# Good
ggplot(dogs, aes(datadog, popularity_all)) + geom_point(color = "blue")

# Bad
ggplot(dogs, aes(datadog, popularity_all)) + geom_point(aes(color = "blue"))
```

You can also use the scales layer to customize the color choices.

Read the documentation for details about parameters.


3. Reverse the y-axis.

Use the scale layer to change axes.

```{r}
ggplot(dogs, aes(datadog, popularity_all, label = breed)) +
    geom_point(aes(color = group)) +
    geom_text(size = 3, vjust = 0, hjust = 1, nudge_x = -0.05) +
    scale_y_reverse()
```


4. Add titles and labels.

```{r}
ggplot(dogs, aes(datadog, popularity_all, label = breed)) +
    geom_point(aes(color = group)) +
    geom_text(size = 3, vjust = 0, hjust = 1, nudge_x = -0.05) +
    scale_y_reverse() +
    labs(x = "Datadog", y = "AKC Popularity Rank",
         title = "Best In Show", color = "Dog Type")
```

Could use ggimage package to replace the points with images
of dogs.


Factors
=======

R represents categorical data using the class `factor`:
```{r}
str(dogs)

class(dogs$group)
```

The categories of a factor are called "levels". You can list the
levels with the `levels()` function:
```{r}
dogs$group

levels(dogs$group)
```


Factors remember all possible levels even if you take a subset:
```{r}
library(dplyr)

first(dogs$group)

head(dogs$group, 1)
```
This is one of the ways factors are different from strings.

You can make a factor forget levels that aren't present with
`droplevels()`:
```{r}
subset = head(dogs$group, 1)
droplevels(subset)
```

_The lecture ended here. I've included the notes below for completeness._

---


You can create a factor with the `factor()` function:
```{r}
factor(c("a", "b", "a"))
```


Internally, R represents factors as integer vectors:
```{r}
# unclass() returns R's internal representation of an object.
unclass(dogs$group)
```
Each integer corresponds to one level of the factor.

This representation uses less memory than repeating each level's
name.
