---
title: "Stat 33A - Lecture 10"
date: April 20, 2020
output: pdf_document
---

Loops
=====


For-loops
---------

```{r}
message(1)
message(2)
message(3)
```

A for-loop iterates over a vector or list:
```{r}
for (i in 1:10) {
  message(i)
  message("Hello")
}

for (i in 1:10)
  message(i)
message("Hello")

x = 1:5
for (xvalue in x) {
  message(xvalue)
}

x = 1:10
for (i in x) {
  i^2
}
```

Use `next` to immediately skip to the next iteration:
```{r}
# Print even numbers from 1 to 10

for (i in 1:10) {
  if (i %% 2 == 0)
    message(i)
}

for (i in 1:10) {
  message("Hello")
  
  if (i %% 2 == 1)
    next
  
  message(i)
}
```

Use `break` to immediately exit the loop:
```{r}
# Find the first number divisible by 7:

for (i in 20:30) {
  message(i)
  
  if (i %% 7 == 0)
    break
  
}
```


Preallocation
-------------

These functions allocate vectors:

* `character()`
* `complex()`
* `numeric()`
* `logical()`

* `vector()`
* `rep()`

```{r}
results = character(5)
complex(3)
numeric(10)
logical(6)

vector("logical", 9)
vector("list", 3)

rep(NA, 4)

rep(-1, 5)

rep("hi", 3)

rep(c("hello", "Hi"), 4)
```


The bad way to store results from a loop:
```{r}
results = c()
for (i in 1:10) {
  results = c(results, i^2)
}
```

Preallocate!

If you know how many results a loop will produce, allocate the result
vector BEFORE the loop:
```{r}
n = 1e5

# BAD:
results = c()
for (i in 1:n) {
  results = c(results, i^2)
}

# GOOD:
results = numeric(n)
for (i in 1:n) {
  results[i] = i^2
}

x = c(-5, 6, 1, 3, 2.2)
n = length(x)
results = numeric(n)
for (i in 1:n) {
  message("New Iteration")
  message(i)
  message(x[i])
  results[i] = x[i]^2
}
```


Use `seq_along(x)` instead of `1:length(x)`:
```{r}
x
results = numeric(length(x))
for (i in seq_along(x)) {
  results[i] = sin(x[i])
}

sin(x)
x^2
```

Iterations can be dependent:
```{r}
# Cumulative sums of a vector.
y = c(1, -3, 2, 3, 4)
#     1  -2  0  3  7
results = numeric(length(y))
for (i in seq_along(y)) {
  if (i == 1) {
    results[i] = y[i]
  } else {
    results[i] = y[i] + results[i - 1]
  }
}
```


While-loops
-----------

A while-loop iterates while some condition is `TRUE`:
```{r}
i = 1
while (i < 11) {
  message(i)
  i = i + 1
}

for (i in 1:10) {
  message(i)
}

```

The `next` and `break` commands also apply to while-loops.



Recursion
=========

A recursive function is one that calls itself:
```{r}
fact = function(n) {
  if (n == 1)
    return (1)
  
  n * fact(n - 1) # n! = n (n - 1)!
}
```

Also see `?Recall`.



Additional Examples
===================

```{r}
x = c(3, 6, 2, 4)

# First iteration code:
message(x[1])

# Edit to iterate more than once:
for (i in seq_along(x)) {
  message(x[i])
}
```
