---
title: "Stat 33B - Lecture 11"
date: April 8, 2020
output: pdf_document
---


Messages, Warnings, & Errors
============================

The `message()` function prints message to the R console:
```{r}
message("Hello everyone!")

message("Hello everyone!", " This is my message.")

message("Hello everyone!", 5)

message("Hello\nHi")

x = message("Hi")

x
```


The `warning()` function prints a warning:
```{r}
if (c(TRUE, FALSE)) {
  5
}

warning("This is your final warning!")
```

The argument to `call.` controls whether the name of the function
that issued the warning is printed:

```{r}
f = function() warning("This is the function warning!", call. = FALSE)

f()
```


The `stop()` function prints an error message and stops evaluation:
```{r}
stop("This is an error")

f = function() stop("This is an error in a function!", call. = FALSE)

f()
```
The `stop()` function also has a `call.` parameter.


These three functions -- `message()`, `warning()`, and `stop()` --
are part of R's "condition system".

The condition system provides a way to report and handle unusual
conditions.


Handling Unusual Conditions
---------------------------

Use the `try()` function to try running an expression that might
produce an error:
```{r}
x = "hi"
result = try(5 + x)

class(result)

if (inherits(result, "try-error")) {
  # Do something about the error
} else {
  # No error, proceed as normal
}
result
```

If the expression succeeds, `try()` returns the result.


If the expression fails (an error), `try()` returns an object with
class `try-error`.

The error can also be silenced by setting `silent = TRUE`:
```{r}
result = try(5 + "hi", silent = TRUE)
result
```

The `tryCatch()` function provides another way to handle conditions.

The function takes "handlers" for different conditions:

* `message`
* `warning`
* `error`
* `interrupt` -- when the user interrupts the code (for example, by
  pressing Ctrl-C)

Each handler should be a function with one parameter. If the
condition occurs, the handler is called with the condition object.

The `tryCatch()` function returns the result of the expression if no
unusual conditions occurred. Otherwise, it returns the result of
calling the handler.

For example:
```{r}
x = "Hi"
result = tryCatch(5 + x,
  error = function(e) {
    warning("x wasn't compatible with 5")
    NA
  })


f = function() warning("This is a very bad warning!")
result = tryCatch(f(),
  warning = function(e) {
    NA
  })

```

With `tryCatch()`, handled conditions are always silent.



The `cat()` Function
------------------

Like `message()`, the `cat()` function prints a message to the
console:
```{r}
cat("Hello", "Hi")
```

But `cat()` is not part of R's condition system.

Messages printed with `cat()` cannot be handled with `tryCatch()`.


The `print()` Function
--------------------

The `print()` function prints R's representation of an object to the
console.

The output is numbered:
```{r}
print("Hello\nHi")
```

Note that strings keep their quotes and special characters are not
converted.


For most output, use the condition system rather than `cat()` or
`print()`!


Formatting Output
-----------------

The `paste()` function combines strings:
```{r}
x = paste("Hi", "Hello")
x
```

The `sprintf()` function can substitute values into strings.

Substitutions are marked with `%` followed by a character:
```{r}
x = 5
sprintf("The value of x is %i. And the next value is %s", x, "Hello")
```

See `?sprintf` for all the details.

The `sprintf()` function may be familiar to you from other languages.



Global Options
==============

The `options()` function sets global options for the R session.

Options have many uses (see `?options`).

One is controlling errors and warnings.



Example: the `warn` option controls how warnings R displayed.

To make R treat warnings as errors:
```{r}
options(warn = 2)
```
This is useful for finding the source of a warning.


The `warn` option can be set to three different values:

* `warn = 0` (the default) warnings are only displayed after code
  finishes running

* `warn = 1` warnings are displayed immediately

* `warn = 2` warnings stop code from running, like errors



It's possible to override the default options every time R starts.

At startup, R runs code from an `.Rprofile` file, usually located in
your system's home directory.

```{r}
normalizePath("~")
```


See `?Startup` for full details.



A few other useful options to make R warn about partial name matches
in calls and the `$` operator:

* `warnPartialMatchAttr = TRUE`

* `warnPartialMatchDollar = TRUE`

* `warnPartialMatchArgs = TRUE`

For example:
```{r}
x = list(abc = 1)
x$a
```


Preventing Bugs
===============

Code Clarity
------------

Format your code so that it's easy to read:

* Use whitespace:

    + Put spaces after commas and around binary operators.
    + Indent code inside curly braces `{ }` by 2 spaces or 1 tab.
    + Don't mix tabs and spaces.
    + Separate logical steps or "paragraphs" with blank lines.

* Put closing curly braces `}` on their own line.
    + Exception: use `} else {`

* Use a consistent naming style:
   + `lowercase_with_underscores`
   + `camelCase`

* Use descriptive variable names. Short names are okay for
  frequently-used variables where the context makes the meaning
  apparent.

* Use comments:
    + To create a big picture plan for how to write your code.
    + To explain tricky code.
    + To summarize the purpose of a "paragraph" of code.
    + To document how to use your functions. Also see roxygen2.


Without formatting:
```{r}
f=function(x,xx){
xx=match.arg(xx,c("celsius","fahrenheit"),several.ok=TRUE)
x1=xx=="fahrenheit"
x[x1]=(x[x1]-32)*5/9
x+273.15}
```

With formatting:
```{r}
to_kelvin = function(temperature, unit) {
  unit = match.arg(unit, c("celsius", "fahrenheit"),
    several.ok = TRUE)

  # Convert everything to celsius first.
  is_f = unit == "fahrenheit"
  temperature[is_f] = (temperature[is_f] - 32) * 5 / 9

  temperature + 273.15
}
```

Modularity
----------

Try to break problems into smaller steps.

Write down the input(s) and output(s) for each step.

Create functions for steps that are reusable. Use parameters for
input and the return value for output.

Be wary of extremely short (1 line) or long (> 20 line) functions.
Often a sign of failure to break down the problem.


Defensive Programming
---------------------

Test whether inputs satisfy your assumptions.

For instance, if your function should only work on scalars, check the
lengths of the arguments.

Raise an error with `stop()` if the assumptions aren't satisfied.

Useful functions for testing assumptions:

* `is.X()` functions for testing various properties. For instance,
  `is.character()` to test whether an object is a character vector.

* `inherits()` for testing the class of an object.

* `length()`, `dim()` for testing dimensions.



Debugging
=========

Debugging code is the process of confirming, step-by-step, that what
you belive the code does is what the code actually does.

The key idea is to check each step of the code.

Two strategies:

* Work forward through the code from the beginning.
* Work backward from the source of an error.

R has built-in functions to help with debugging.

The `browser()` function pauses the running code and starts R's
debugging system.

For example:
```{r}
for (i in 1:10) {
  message(i)
  
  if (i == 5)
    browser()
}
```

The key debugger commands are:

* `n` to run the next line
* `s` to "step into" a call
* `c` to continue running the code
* `Q` to quit the debugger
* `where` to print call stack
* `help` to print debugger help


```{r}
f = function() {
  browser()
  g()
  message(4)
}

g = function() {
  message(1)
  message(2)
  message(3)
}

f()
```


The `debug()` function places a call to `browser()` at the beginning
of a function.

Use `debug()` to debug functions that you can't or don't want to
edit.

For example:
```{r}
f = function() {
  message(1)
  message(2)
  message("Hi")
}

debug(f)
```

Use `undebug()` to reverse the effect of `debug()`:
```{r}
undebug(f)
```

The `debugonce()` function places a call to `browser()` at the
beginning of a function for the next call only.

The idea is that you then don't have to call `undebug()`.

Example:
```{r}
debugonce(f)
```

The global option `error` can be used to make R enter the debugger
any time an error occurs.

Set the option to `error = recover`:
```{r}
options(error = recover)
```

Then:
```{r}
f = function() {
  stop("This is an error!")
}

f()
```

See Chapter 13 of "The Art of R Programming" by N. Matloff for more.
