---
title: "Stat 33A - Lecture 7"
date: March 30, 2020
output: pdf_document
---


Reshaping Data
==============

We'll use the tidyr package:
```{r}
# You may need to install the package first:
# install.packages("tidyr")

# Load the package:

```

Most Tidyverse functions requre data in "tidy" form.

The tidyr package provides an example of tidy data:
```{r}

```

Three properties of tidy data:

1. Each observation must have its own row.
2. Each feature must have its own column.
3. Each value must have its own cell.

Convenient for data analysis in general.


An example of untidy data:
```{r}

```


The tidyr package provides functions to "reshape" data into tidy
data.

Usually this means rotating rows into columns, or columns into rows.

Before you reshape, determine:

* Identifiers: values that identify the subject of observation.

* Measurements: values that were measured for the subject of
  observation.

For example:
```{r}

```

The `country` and `year` columns identify the observation.

The `cases` and `population` columns are measurements.



Rotating Rows into Columns
--------------------------

When to rotate rows into columns?

Measurements for one observation are split across several rows.

We saw this earlier:
```{r}

```

To make this data tidy:

* Rotate the `count` row values into columns, by `type`.
* New columns are `cases` and `population`.

Tidying this data set makes it wider.


Use `pivot_wider()` to rotate rows into columns:
```{r}

```

Need to specify:

* Column names to rotate as `names_from`.
* Measurements to rotate as `values_from`.


Rotating Columns into Rows
--------------------------

When to rotate columns into rows?

Each row contains measurements for multiple observations.

In this table, each row contains the measurements for two
observations of the `cases` variable:
```{r}

```

To make this data tidy:

* Rotate the `1999` and `2000` columns into rows.
* New columns are `year` and `cases`.

Tidying this data set makes it longer.


Use `pivot_longer()` to rotate columns into rows:
```{r}

```

Need to specify:

* Columns to rotate as `cols`.
* Name(s) of new identifier column(s) as `names_to`.
* Name(s) of new measuerment column(s) as `values_to`.


See `vignette("pivot")` for more examples of using `tidyr`.


Relational Data
===============

Data split across multiple tables are called _relational data_.

A column shared by several tables is called a _key_.

For example, a grocery store's inventory system might have:

* A table that lists stores
* A table that lists items (fruits, vegetables, etc)
* A table that lists quantity of each item at each store

The data set:
```{r}

```

A "join" combines two separate tables by based on a common key.

For each row in the first table, the key is compared against the rows
in the second table.

If the key matches, the rows are combined into a new row.

Note that a join can duplicate rows!



Joins with dplyr
----------------

The dplyr package provides several join functions.

The `inner_join()` function only keeps rows where the key matched:
```{r}

```

Use the `by` parameter to specify the key:
```{r}

```


The `full_join()` function keeps all rows, regardless of whether the
key matched.

As a result, some rows will contain missing values:
```{r}

```

The `left_join()` function keeps all rows in the "left" or first
table.

Rows in the right table are only kept if they match:
```{r}

```

Conversely, the `right_join()` function keeps all rows in the "right"
or second table.

```{r}

```

So the four kinds of joins are:

* Inner: keep matches
* Full: keep everything
* Left: keep all left, keep matching right
* Right: keep matching left, keep all right

Joins are not just used in R!

Especially important to understand for working with databases and
SQL.


The dplyr package also provides other kinds of joins that are not as
common outside of R. See the docs for details.
