---
title: "Stat 33B - Lecture 8"
date: March 11, 2020
output: pdf_document
---

Review: If-statements
=====================

If you use `else`, you need to make sure it's on the same line as
code from the `if` block:
```{r}
x = 4

if (x < 2)  message("Hi") else message("Bye")

y = if (x < 2) "Hi" else "Bye"
y
```
This was a problem I ran into in the previous lecture.


Loops
=====

Four ways to write loops in R:

1. Vectorization (fast!)
2. Apply functions: `lapply()`, `sapply()`, etc
3. `for`, `while`, and `repeat` loops
4. Recursion (slow)


Apply Functions
---------------

Similar to "map" functions in other languages.

Applies a function (FUN) to each element of a vector or list (X):
```
lapply(X, FUN, ...)
```

We saw:

* `lapply()`: always returns a list ("l" for list)
* `sapply()`: attempts to return a vector ("s" for simplify)
* `vapply()`: always returns a vector ("v" for vector)


`mapply()` applies a function to each element of multiple vectors or
lists:
```{r}
data1 = list(1:3, 5:8)
data2 = list(3)

sum(data1[[1]], data2[[1]])
sum(data1[[2]], data2[[2]])

mapply(sum, data1, data2)
```
"m" for **multiple** vectors.


### Flashback to Data Analysis

The `split()` function splits a vector or data frame into groups
based on some other vector.

```{r}
dogs = readRDS("data/dogs/dogs_full.rds")

split(dogs, dogs$kids)

split(dogs$weight, dogs$group)
```

The `split()` function is especially useful when combined with
`lapply()` or `sapply`().

```{r}
by_group = split(dogs$weight, dogs$group)
sapply(by_group, median, na.rm = TRUE)
```
This is an R idiom!


The `tapply()` function is equivalent to the `split()` and `sapply()`
idiom:
```{r}
tapply(dogs$weight, dogs$group, median, na.rm = TRUE)
```
"t" for **table**, because `tapply()` is a generalization of the
frequency-counting function `table()`.


The `split()` and `tapply()` functions are important for analyzing
data in data frames.


### Even More Apply Functions

* `apply()` applies a function to rows, columns, or elements of a
  matrix, and returns a matrix.
* `eapply()` applies a function to the elements of an environment.
  We'll learn about environments in a later lecture.
* `rapply()` applies a function to all elements of a recursive list
  (a list of lists).

See this StackOverflow Post for a good summary:

    https://stackoverflow.com/a/7141669

The Tidyverse package purrr provides alternatives to R's apply
functions.


`for`, `while`, and `repeat`
----------------------

A for-loop runs a block of code once for each element of a vector or
list.

```{r}
for (i in 1:10) {
  message(i)
}

x = lapply(1:10, message)

# Fibonacci: 1 1 2 3 5 8 13 ...

n = 15
fib = c(1, 1)
for (i in 3:n) {
  fib[i] = fib[i - 1] + fib[i - 2]
}
```
The same idea as for-loops in other languages.


A while-loop runs a block of code repeatedly as long as some
condition is true:
```{r}
n = 15
fib = c(1, 1)
i = 3
while (i <= n) {
  fib[i] = fib[i - 1] + fib[i - 2]
  i = i + 1
}
```
Again, the same idea as in other languages.


Use `break` to exit a loop early:
```{r}
for (i in 1:10) {
  if (i %% 3 == 0) {
    message("Stopped")
    message(i)
    break
  }
}
```

Use `next` to skip to the next iteration early:
```{r}
for (i in 1:10) {
  if (i %% 3 == 0) {
    message("Div 3")
    next
  }
  message(i)
}
```


A repeat-loop runs a block of code repeatedly:
```{r}
n = 15
fib = c(1, 1)
i = 2
repeat {
  i = i + 1
  fib[i] = fib[i - 1] + fib[i - 2]
  if (i >= n)
    break
}
```

If you know how many results you expect to get from the loop,
allocate storage before the loop begins:
```{r}
n = 15
fib = numeric(n)
fib[1:2] = 1
for (i in 3:n) {
  fib[i] = fib[i - 1] + fib[i - 2]
}
```

If you need indices, using `1:length(x)` can cause bugs:
```{r}
1:0

x = numeric(0)
1:length(x)

for (i in seq_along(x)) {
  message(x[i])
}
```

Use `seq_along(x)` instead.


Recursion
---------

A recursive function is one that calls itself.

We'll see an example of this the next section.


How to Write Loops
------------------

When thinking about writing a loop, try (in order):

1. vectorization (fast!)
2. apply functions
    * Try an apply function if iterations are independent.
3. for/while-loops
    * Try a for-loop if some iterations depend on others.
    * Try a while-loop if the number of iterations is unknown.
4. recursion
    * Convenient for naturally recursive problems (like Fibonacci),
      but often there are faster solutions.

Before you write the loop, try writing the code for just 1 iteration.

Make sure that code works; it's easy to test code for 1 iteration.

When you have 1 iteration working, then try using the code in a loop
(you will have to make some small changes).

If your loop doesn't work, try to figure out which iteration is
causing the problem. One way to do this is to use message() to print
out information.

Then try to write the code for the broken iteration, get that
iteration working, and repeat this whole process.


Functions
=========

You can write your own function with the `function` command.

Generally, you'll assign the function to a variable:
```{r}
hi = function() {
  message("Hi")
}

hi()
```

Use curly braces `{ }` to write a function with multiple lines of
code.

The last expression in the function is returned:
```{r}
square = function(x) x^2

square(7)

square(c(5, 6, 7))

cube = function(x) {
  x^3
}

cube(c(-1, 1, 2))
```

Use `return()` to return a result immediately (exiting the function):
```{r}
cube0 = function(x) {
  if (x == 0)
    return (42)
  
  x^3
}

cube0(0)
```
It's idiomatic to only use `return()` when strictly necessary.


Parameters can have default arguments that are used if no argument is
provided.

Use `=` to assign default arguments:
```{r}
cube0 = function(x = -1) {
  if (x == 0)
    return (42)
  
  x^3
}

cube0()
```


Example: Recursion
------------------

Recall that a recursive function is one that calls itself:
```{r}
# Compute nth element of Fibonacci
find_fib = function(n = 15) {
  if (n %in% c(1, 2))
    return(1)
  
  find_fib(n - 1) + find_fib(n - 2)
}

find_fib(15)
```
This creates a kind of loop.


R's built-in `Recall()` function will re-call the enclosing function:
```{r}
foo = function(n = 15) {
  if (n %in% c(1, 2))
    return(1)
  
  Recall(n - 1) + Recall(n - 2)
}

foo(15)
```

Without `Recall()`, renaming a recursive function can break the
function.


Lazy Evaluation
---------------

Arguments to a function are only evaluated if they're used:
```{r}
f = function(x) 42

f(message("Hi!"))
```
This is called "lazy evaluation".


You can force evaluation with the `force()` function:
```{r}
f = function(x) {
  force(x)
  42
}

f(message("Hi!"))
```


Because of lazy evaluation, you can use one parameter as part of a default
argument to another parameter:
```{r}
add = function(x, y = x) {
  x + y
}

add(2)

add(2, 3)
```

Default arguments can even be circular:
```{r}
add = function(x = y, y = x) {
  x + y
}

add(4)

add(y = 3)

add()
```


The Dots Parameter
------------------

The dots parameter `...` can accept any number of arguments.

See `?dots` for documentation on the dots parameter.

The dots parameter is not a normal R object. You need to use special
functions to handle it.

The `...length()` function returns the number of arguments to the
dots parameter:
```{r}
count_args = function(...) {
  ...length()
}

count_args()

count_args(1, 4)

count_args("hi")
```

The `...elt()` function gets the nth argument to the dots parameter:
```{r}
get_nth = function(n, ...) {
  ...elt(n)
}

get_nth(4, "a", "b", "c")
```

You can also pass arguments to the dots parameter to another
function:
```{r}
trace_mean = function(x, trace = 0.01, ...) {
  mean(x[x > trace], ...)
}

trace_mean(c(1, 2, 0.005, NA), na.rm = TRUE)
```



How to Write Functions
----------------------

Functions are the building blocks for solving problems.

Take a divide-and-conquer approach, breaking large problems into
smaller steps.

Use a short function for each step. Then it's easier to:

* Test that each step works correctly.
* Modify, reuse, or repurpose a step.

Before you write a function, it's useful to go through several steps:

1. Write down what you want to do, in detail. It can also help to
   draw a picture of what needs to happen.

2. Check whether there's already a built-in function. Google it!

3. Write the code to handle a simple case first. For data science
   problems, use a small dataset at this step.
