---
title: "Stat 33A - Lecture 11"
date: April 27, 2020
output: pdf_document
---

Code Style & Organization
=========================


Some code with unclear style and organization:
```{r}
f=function(x,xx)
{if(!all(xx%in%c("celsius","fahrenheit")))
stop("invalid units")
x1=xx=="fahrenheit"
x[x1]=(x[x1]-32)*5/9
x+273.15}
```


A clearer version:
```{r}

```


Preventing Bugs
===============

Defensive Programming
---------------------


Useful functions for testing assumptions:

* `is()` to check data type

* `length()` and `dim()` to check dimensions

* `stop()` to print an error message and stop evalutation


Examples:
```{r}

```

R Pitfalls
----------

Advice based on what I've seen this semester.


Don't mix up `==` and `%in%`:
```{r}

```

Don't use `== TRUE`:
```{r}

```

Don't use `== FALSE` to negate conditions:
```{r}

```


Don't use `if (condition) TRUE else FALSE`:
```{r}

```

Don't use `print()` to print messages:
```{r}

```

Don't forget `library()`.





Debugging
=========

The `browser()` function pauses the running code and starts R's
interactive debugging system.

For example:
```{r}

```

The key debugger commands are:

* `n` to run the next line
* `s` to "step into" a call
* `c` to continue running the code
* `Q` to quit the debugger
* `where` to print the call stack
* `help` to print debugger help


```{r}

```
