---
title: "Stat 33A - Lecture 11"
date: April 27, 2020
output: pdf_document
---

Code Style & Organization
=========================


Some code with unclear style and organization:
```{r}
f=function(x,xx)
{if(!all(xx%in%c("celsius","fahrenheit")))
stop("invalid units")
x1=xx=="fahrenheit"
x[x1]=(x[x1]-32)*5/9
x+273.15}
```


A clearer version:
```{r}
to_kelvin = function(temp, unit) {
  # Check unit because this function only handles Celsius and Fahrenheit.
  if (!all(unit %in% c("celsius", "fahrenheit")))
    stop("elements of unit must be celsius or fahrenheit")
  
  # Convert Fahrenheit to Celsius, then Celsius to Kelvin
  is_fahren = unit == "fahrenheit"
  temp[is_fahren] = (temp[is_fahren] - 32) * 5 / 9
  temp + 273.15
}
```


Preventing Bugs
===============

Defensive Programming
---------------------


Useful functions for testing assumptions:

* `is()` to check data type

* `length()` and `dim()` to check dimensions

* `stop()` to print an error message and stop evalutation


Examples:
```{r}
is(5, "character")

is(5, "numeric")

x = "hello"
is(x, "logical")
```

R Pitfalls
----------

Advice based on what I've seen this semester.


Don't mix up `==` and `%in%`:
```{r}
x = c(1, 2, 3)
# Which are equal to 2 or 3?

x %in% c(3, 2)

x == c(3, 2)
# 1 2 3
# 3 2 3
# F T T

x == c(1, 3, 3)
x == 5
```

Don't use `== TRUE`:
```{r}
x = c(1, 2, 3)

x > 1
```

Don't use `== FALSE` to negate conditions:
```{r}
# BAD:
(x < 1) == FALSE

# GOOD:
!(x < 1)

x >= 1
```


Don't use `if (condition) TRUE else FALSE`:
```{r}
x = "hi"

# BAD:
if (x == "hello") TRUE else FALSE

if (x == "hello") 1 else 0

# GOOD:
x == "hello"

# TRUE is 1, FALSE is 0
as.integer(x == "hello")
```

Don't use `print()` to print messages:
```{r}
# BAD:
print("Hello\nClass")

# GOOD:
message("Hello\nClass")
```

Don't forget `library()`.





Debugging
=========

The `browser()` function pauses the running code and starts R's
interactive debugging system.

For example:
```{r}
for (i in 1:10) {
  browser()
  message(i)
}
```

The key debugger commands are:

* `n` to run the next line
* `s` to "step into" a call
* `c` to continue running the code
* `Q` to quit the debugger
* `where` to print the call stack
* `help` to print debugger help


```{r}
f = function(msg) {
  message("Hello")
  browser()
  g()
  message(msg)
}

g = function() {
  message("The call to g() succeeded!")
}

f("Goodbye")
```
