---
title: "Stat 33B - Lecture 5"
date: February 19, 2020
output: pdf_document
---

Announcements
=============

Grades posted (please check):

* Homework 1
* Lab 1
* Quiz 1
* Attendance up to 12 Feb

Solutions are posted!


Homework 3 will be posted later today or tomorrow.


Previous Lecture: Visualizations
================================

We started working with the dogs data to recreate the Best in Show
visualization:

    https://informationisbeautiful.net/visualizations/
        best-in-show-whats-the-top-data-dog/


```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
```

In `ggplot2`, graphics are composed of layers.

Layer 1: Data -- the data frame to plot.
Layer 2: GEOMetry -- shapes to represent data.
Layer 3: AESthetic -- "wires" between geometry and data


```{r}
library(ggplot2) 

dogs[ , "datadog"]
dogs[["datadog"]]

# Aesthetic layer applies to every geometry layer:
ggplot(dogs, aes(x = datadog, y = popularity)) + geom_point()

# Alternate syntax:
# Aesthetic layer only applies to geom_point:
ggplot(dogs) + geom_point(aes(x = datadog, y = popularity))
```



More About `ggplot2`
==================

Important layers in `ggplot2`:

Layer       | Description
----------  | -----------
data        | A data frame to visualize
aesthetics  | The map or "wires" between data and geometry
geometry    | Geometry to represent the data visually
statistics  | An alternative to geometry
scales      | How numbers in data are converted to numbers on screen
labels      | Titles and axis labels
guides      | Legend settings
annotations | Additional geoms that are not mapped to data
coordinates | Coordinate systems (Cartesian, logarithmic, polar)
facets      | Side-by-side panels


Saving Your Plots
-----------------

Use `ggsave()` to save the last plot you created to a file:

```{r}
ggsave("my33bplot1.png")
```

__Important Note:__ RStudio can maintain several different working directories
at the same time. The working directory for code run from a notebook might be
different from code entered manually in the R prompt! This is the reason my
plot showed up in an unexpected place during this lecture.

```{r}
ggsave("~/university/teach/stat33ab/stat33b/sandbox/myplot.jpeg")
```



Adding More Details
-------------------

How else can we make our plot look more like the Best in Show plot?


1. Add the dog breeds as text.

Add more geometries to add additional details to a plot:
```{r}
ggplot(dogs, aes(x = datadog, y = popularity))  +
  geom_point() +
  geom_text(aes(label = breed), size = 2, hjust = "left",
            vjust = "bottom", nudge_x = 0.05)
```

See the `ggrepel` package for automatic label positioning.


2. Color the points by type of dog.

```{r}
ggplot(dogs, aes(x = datadog, y = popularity))  +
  geom_point(aes(color = group)) +
  geom_text(aes(label = breed), size = 2, hjust = "left",
            vjust = "bottom", nudge_x = 0.05)
```

Where you set the aesthetics matters. In the code above, the `x` and `y`
aesthetics apply to ALL geometries. The `color` aesthetic only applies to
`geom_point`, and the `label` aesthetic only applies to `geom_text`. To make
these aesthetics apply to everything, move them to the `aes` call in the data
layer.


We can also set parameters outside of the aesthetics.
```{r}
ggplot(dogs, aes(x = datadog, y = popularity))  +
  geom_point(color = "purple")
```

Doing so sets a constant value instead of mapping to a feature in the
data.

You can also use the scales layer to customize the color choices.

Read the documentation for details about parameters.


3. Reverse the y-axis.

Use the scale layer to change axes.

```{r}
ggplot(dogs, aes(x = datadog, y = popularity))  +
  geom_point(aes(color = group)) +
  geom_text(aes(label = breed), size = 2, hjust = "left",
            vjust = "bottom", nudge_x = 0.05) +
  scale_y_reverse()
```


4. Add titles and labels.

```{r}
ggplot(dogs, aes(x = datadog, y = popularity))  +
  geom_point(aes(color = group)) +
  geom_text(aes(label = breed), size = 2, hjust = "left",
            vjust = "bottom", nudge_x = 0.05) +
  scale_y_reverse() +
  labs(title = "Best in Show", x = "Datadog", y = "Popularity")
```



Reshaping Data
==============

The `ggplot2` functions require data in "tidy" form:

1. Each observation must have its own row.
2. Each feature must have its own column.
3. Each value must have its own cell.

Also required by most other Tidyverse packages.

Tidy data sets are convenient for data analysis in general.


The `tidyr` package provides functions to reshape data to and from
tidy form.
```{r}
# You may need to install the package first:
# install.packages("tidyr")

# Load the package:
library(tidyr)
```

An example of tidy data (Tuberculosis Data Set):
```{r}
# Included with tidyr:
table1
```

The `country` and `year` columns identify the observation.

The `cases` and `population` columns are measurements.


Determine the identifiers and the measurements are BEFORE reshaping!



Rotating Rows into Columns
--------------------------

Some data sets split measurements for one observation across several
rows.

For example, `cases` and `population` here:
```{r}
# Included with tidyr:
table2
```

To make this data tidy:

* Rotate the `count` row values into columns, by `type`.
* New columns are `cases` and `population`.

Tidying this data set makes it wider.


Use `pivot_wider()` to rotate rows into columns:
```{r}
pivot_wider(table2, names_from = type, values_from = count)
```

Need to specify:

* Column names to rotate as `names_from`.
* Measurements to rotate as `values_from`.


If you wanted to do this without `tidyr`:

1. Subset rows to separate `cases` and `population` values.
2. Remove the `type` column from each.
3. Rename the `count` column to `cases` and `population`.
4. Merge these two subsets by matching `country` and `year`.

```{r}
# Step 1
cases = table2[table2$type == "cases", ]
pop = table2[table2$type == "population", ]
pop

# Step 2
cases = cases[-3]
pop = pop[-3]

# Step 3
names(pop)[3] = "population"
names(cases)[3] = "cases"

pop
cases

# Risky way to do step 4:
cbind(pop, cases)

# Okay for this data set, because rows line up in country and year.
#
# Other data sets might not have rows that line up.
```

We'll see a better way to do step 4 later.
