---
title: "Stat 33A - Lecture 6"
date: March 16, 2020
output: pdf_document
---


Review
======

Last time:

R represents categorical data using the class `factor`:
```{r}
str(dogs)

class(dogs$group)
```

The categories of a factor are called "levels". You can list the
levels with the `levels()` function:
```{r}
dogs$group

levels(dogs$group)
```

Factors remember all possible levels even if you take a subset:
```{r}
library(dplyr)

first(dogs$group)

head(dogs$group, 1)
```
This is one of the ways factors are different from strings.

You can make a factor forget levels that aren't present with
`droplevels()`:
```{r}
subset = head(dogs$group, 1)
droplevels(subset)
```

You can create a factor with the `factor()` function:
```{r}

```


Internally, R represents factors as integer vectors:
```{r}

```
Each integer corresponds to one level of the factor.

This representation uses less memory than repeating each level's
name.



Exploratory Data Analysis
=========================

What does it mean to "explore" data?

* Look for patterns (examine variation in the data)
* Look for errors in the data
* Look for relationships between variables
* Look at data to get an overview (what data are present?)
* Check assumptions (model, conclusions, etc)

What are the techniques to "explore" data?

* Make plots
* Compute summary statistics
* Fit models (including hypothesis tests, machine learning)


The table below has _suggestions_ for choosing an appropriate plot
based on the data types.

You also need to think about what you're trying to convey.

First Feature  | Second Feature   | Plot
-------------- | ---------------- | ----
categorical    |                  | bar, dot
categorical    | categorical      | bar, dot, mosaic
numerical      |                  | box, density, histogram
numerical      | categorical      | box, density
numerical      | numerical        | line, scatter, smooth scatter



Example: How many dogs are there in each group (toy, sporting, etc)?

```{r}

```

Example: What's the distribution of datadog scores?

```{r}

```

Example: How are size and height related?

```{r}

```


Taking Subsets
==============

We've already seen how to take subsets with dplyr:

* `slice()` to get rows by position or name.
* `filter()` to get rows that satisfy a condition.
* `select()` to get columns by position or name.

Now we'll learn R's built-in functions for taking subsets.

Why learn these?

1. R's built-in functions are more general. They work with data
   frames, lists, vectors, and more.

   The dplyr functions only work with data frames.


2. dplyr is popular, but not everyone uses it.

   If you're going to use R packages or collaborate with people, you
   need to be able to read code written either way.


The fundamental built-in operators for taking subsets are:

1.  The "double square bracket" or "extraction" operator `[[`

    + By position or name.
    + Selects exactly one element.
    + Drops the container.

2. The "square bracket" or "subset" operator `[`

    + By position, name, or logical values.
    + Selects any number of elements.
    + Keeps the container.


The examples will clarify what some of these things mean.


Methods of Selection
--------------------

Let's consider a vector first.
```{r}

```
We'll see examples of data frames later.


Both `[` and `[[` can select elements by position:
```{r}

```

Notice that `[[` cannot select more than one element:
```{r}

```


Negative positions mean "everything except":
```{r}

```


Both `[` and `[[` can select elements by name:
```{r}

```


Only `[` can select elements using a logical vector.

The logical vector should have the same length as the object being
subsetted.

`TRUE` means keep the element in that position, `FALSE` means drop
the element in that position:
```{r}

```

If the length is different, R recycles:
```{r}

```

In R, conditions return logical vectors.
```{r}

```

So `[` can select elements based on a condition:
```{r}

```

A logical vector can contain missing values `NA`.

These are retained in the subset:
```{r}

```


The `which()` function returns the position of each `TRUE` in a
logical vector, ignoring `NA`s:
```{r}

```

So you can use `which()` to drop `NA`s in a condition:
```{r}

```
Do this as needed, not everywhere.

Calling `which()` makes your code a little bit slower.



Containers
----------

In a vector, every element must have the same data type.

That rule does not apply to lists.

Lists are like a container for other data types.

`[` keeps the container, but `[[` drops the container:
```{r}

```

A visual representation of this idea:

    https://twitter.com/hadleywickham/status/643381054758363136


This leads to one oddity with `[`:
```{r}

```


For lists, `[[` does recursive subsetting if you pass a vector of
positions:
```{r}

```

Subsetting Data Frames
----------------------

R thinks of data frames as lists of columns.

So `[` and `[[` get columns:
```{r}

```
`[` keeps the data frame, `[[` drops the data frame.


Data frames are 2-dimensional.

Both `[` and `[[` can subset each dimension separately. The syntax
is:

    DATA[ROWS, COLUMNS]

Use a blank to mean "everything" along one dimension.

Examples:
```{r}


```

Finally, `[` will drop the container if you only select one column:
```{r}

```

You can control this behavior with the `drop` parameter:
```{r}

```
The default is `drop = TRUE`.


One exception!

Data frames with class "tbl" and "tbl_df" are called tibbles, and
part of a Tidyverse package called tibble.

Notice that the dogs data frame is a tibble:
```{r}

```

For tibbles, the default is `drop = FALSE`.

This only applies if you load a Tidyverse package:

  * tibble
  * dplyr
  * ggplot2
  * ...

Then using `[` to subset a single value DOES NOT drop the data frame:
```{r}

```


More Ways to Take Subsets
-------------------------

Two more built-in functions for taking subsets:

1. `$` extracts an element by name. Tries to match partial names.
2. `subset()` takes subsets of rows in a data frame.

Most useful in interactive programming!

The `$` is similar to `[[`:
```{r}

```

Use quotes with `$` when names contain characters that are R
operators (such as `+`, `-`, `>`, ...):
```{r}

```

Unlike `[[`, the `$` will try to match partial names:
```{r}

```

The `subset()` function is a shortcut to avoid typing the name of the
data frame over and over when subsetting rows with a condition.

For example:
```{r}

```

Equivalent, using `subset()`:
```{r}

```
With `subset()`, there's no need to use `$`, and the `NA`s are
eliminated automatically.
