---
title: "Stat 33B - Lecture 4"
date: February 12, 2020
output: pdf_document
---

Announcements
=============


Working with Data
=================

We'll need the dogs data again:
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
```

Taking Subsets
--------------

Taking subsets is a fundamental operation for working with data in R!

Two primary operators for taking subsets in R:

1.  `[[` to "extract" an element.

    + Selects exactly one element.
    + Drops the container.
    + By position or name.

2. `[` to "subset" an element.

    + Selects any number of elements.
    + Keeps the container.
    + By position, name, or logical values.


Let's test these out on a vector:
```{r}
x = c(7, -3, 1)
```

Subsetting by position (positions start at 1 in R):
```{r}
x[1]

x[[1]]

x[c(1, 2, 2)]

x[[c(1, 2, 2)]]
```

Negative positions mean "everything except":
```{r}
x[-2]

x[[-1]] # can't return multiple elements
```


Subsetting by name:
```{r}
x
y = c(a = 7, b = -5, c = 1)
y

y["a"]

x["a"] # no element named "a"

y[["b"]]

x[["b"]] # error

y[c("a", "b", "a")]
```


Subsetting by logical value:
```{r}
y > 0

which(y > 0)

y[which(y > 0)] # subsetting by position

y[y > 0]

# One case where this breaks:
y_withNA = c(7, -1, NA)

y_withNA > 0

y_withNA[y_withNA > 0]

y_withNA[which(y_withNA > 0)] # drops the NAs
```

Using `which()` means an extra call, so extra time!

```{r}
x[[c(FALSE, FALSE, TRUE)]]
```

```{r}
# If length of logical doesn't match, R recycles:
x

# FALSE TRUE FALSE TRUE FALSE TRUE ...
x[c(FALSE, TRUE)]
```

`[[` drops the container, `[` does not.

This is most apparent for lists:

```{r}
x = list(a = c(6, 5), b = "hi")
x

class(x[1])

class(x[[1]])

x

x[1]

x[[1]]

x[[c(2, 2)]]

x[[2]]

x[[c(1, 2)]] # recursive subsetting: 1st element's 2nd element

x[[1]][[2]] # equivalent to above line

x[[1]]

x[1][1][1][1] # keeps the container

x[c(1, 1, 2)]
```

A visual representation of this idea:

    https://twitter.com/hadleywickham/status/643381054758363136


Data frames are lists, so:
```{r}
class(dogs[1])

dogs[[1]]
```

We can also do multi-dimensional subsetting:
```{r}
head(dogs)

dogs[1, 1] # drops the container

# NOTE: Data frames with class "tbl" and "tbl_df" are called tibbles, and part
# of a Tidyverse package called tibble.
#
# Notice that the dogs data frame is a tibble.
#
# In recent versions of the tibble package, using `[` to subset a single value
# DOES NOT drop the data frame.
#
# So you may see different behavior above if you have a recent version of the
# tibble package.
#
# If you want to see the behavior for ordinary data frames, try subsetting a
# non-tibble data frame, such as R's built-in `iris` data frame:
iris[1, 1]

# Thanks to the student that pointed out the behavior of `[` is now different
# for tibbles.
#
# You can also force R to drop the data frame with the `drop` parameter:
dogs[1, 1, drop = TRUE]


# The `[[` operator ALWAYS drops the container:
dogs[[1, 1]]


# If you select multiple elements, `[` always keeps the container:
dogs[1, 1:3] # row 1, columns 1 through 3
```

Use a blank to mean "everything" along one dimension:
```{r}

dogs[1, ] # first row, all columns
```


Two more functions for taking subsets:

1. `$` extracts an element by name. Tries to match partial names.
2. `subset()` takes subsets of row in a data frame.

Most useful in interactive programming!

The `$` is similar to `[[`:
```{r}
x[["a"]]

x$"a"
```

Use quotes with `$` when names contain characters that are R operators (such as
`+`, `-`, `>`, ...):
```{r}
x$"+A"
```

Unlike `[[`, the `$` will try to match partial names:
```{r}
dogs$bree # breed
```

The `subset()` function is a shortcut to avoid typing the name of the data
frame over and over when subsetting rows with a condition.

For example:
```{r}
head(dogs)

colnames(dogs)

above = dogs$weight > mean(dogs$weight, na.rm = TRUE)

# which() to eliminate NAs
dogs[which(above), ]
```

Equivalent, using `subset()`:
```{r}
subset(dogs, weight > mean(weight, na.rm = TRUE))
```
With `subset()`, there's no need to use `$`, and the `NA`s are eliminated
automatically.


Transforming Data Frames
------------------------

A few more functions for working with data frames:

* `rbind()` stacks two or more data frames on top of each other.
* `cbind()` combines two or more data frames side-by-side.
* `t()` transposes a data frame (or matrix).

Examples:
```{r}
dogs[1:3, ]

rbind(dogs[1:3, ], dogs[1:3, ])
```

These functions also work with matrices.


Creating Visualizations
=======================

There are three main systems for creating visualizations in R:

1. The base R functions, including `plot()`.

2. The `lattice` package. The interface is similar to the base R
   functions, but uses lists of parameters to control plot details.

3. The `ggplot2` package. The interface is a "grammar of graphics"
   where plots are assembled from layers.
  
Both `lattice` and `ggplot2` are based on R's low-level `grid`
graphics system.

It is usually easier to customize visualizations made with base R.

Both `lattice` and `ggplot2` are better at handling grouped data and
generally require less code to create a nice-looking visualization.

We'll learn `ggplot2`.


R Packages and CRAN
-------------------

The Comprehensive R Archive Network (CRAN) is a repository of
user-contributed packages for R.

You can install packages from CRAN with `install.packages()`.

For example:
```{r}
install.packages("ggplot2")
```

For maintaining your packages, there are also functions:

* `remove.packages()` to remove a package
* `update.packages()` to update ALL packages
* `installed.packages()` to list installed packages

You can load an installed package into your R session with the
`library()` function.
```{r}
library(ggplot2)
```

Install once, load with `library()` each time you restart R.


The Tidyverse
-------------

The Tidyverse (<https://www.tidyverse.org/>) is collection of R
packages for doing data science in R.

They are made by many of the same people that make RStudio, and
provide alternatives to R's built-in tools for:

* Reading files (package `readr`)
* Manipulating data frames (packages `dplyr`, `tidyr`, `tibble`)
* Manipulating strings (package `stringr`)
* Manipulating factors (package `forcats`)
* Functional programming (package `purrr`)
* Making visualizations (package `ggplot2`)

The Tidyverse packages are popular but controversial, because some of
them use a syntax different from base R.

RStudio cheat sheets (mostly for Tidyverse packages):

    https://rstudio.com/resources/cheatsheets/


Making Graphics with ggplot2
----------------------------

The fundamental idea of `ggplot2` is that all graphics are composed
of layers.


Documentation:

    https://ggplot2.tidyverse.org/


As an example, let's create a simplifed version of the Best in Show
visualization:

    https://informationisbeautiful.net/visualizations/
        best-in-show-whats-the-top-data-dog/


Layer 1: Data -- the data frame to plot.

Call the `ggplot()` function to set the data layer:
```{r}
library(ggplot2)

ggplot(dogs)
```

Layer 2: GEOMetry -- shapes to represent data.

Add a `geom_` function to set the geometry layer:
```{r}
ggplot(dogs) + geom_point()
```

Layer 3: AESthetic -- "wires" between geometry and data

Call the `aes()` function in `ggplot()` or the `geom_` to set the
mapping between the data layer and geometry layer:
```{r}
ggplot(dogs, aes(x = datadog, y = popularity)) + geom_point()

# Alternative syntax:
ggplot(dogs) + geom_point(aes(x = datadog, y = popularity))
```

Important layers in `ggplot2`:

Layer       | Description
----------  | -----------
data        | A data frame to visualize
aesthetics  | The map or "wires" between data and geometry
geometry    | Geometry to represent the data visually
statistics  | An alternative to geometry
scales      | How numbers in data are converted to numbers on screen
labels      | Titles and axis labels
guides      | Legend settings
annotations | Additional geoms that are not mapped to data
coordinates | Coordinate systems (Cartesian, logarithmic, polar)
facets      | Side-by-side panels


Saving Your Plots
-----------------

Use `ggsave()` to save the last plot you created to a file:

```{r}
ggsave("myplot.png")
```
