---
title: "Stat 33A - Lecture 9"
date: April 13, 2020
output: pdf_document
---


Review: Vectorization
=====================

A "vectorized" function is one that automatically applies an
operation to each element of a vector.

We saw this earlier with arithmetic:
```{r}

```

Important takeaway: vectorized functions provide a way to do the same
operation repeatedly.



Apply Functions
===============

Doing the same operation repeatedly is a common pattern in
programming.

Vectorization is one way, but not all functions are vectorized.

In R, the "apply functions" are another way to do something
repeatedly.

The apply functions call a function on each element of a vector or
list.


The `lapply()` Function
---------------------

The first and most important apply function is `lapply()`. The syntax
is:
```
lapply(X, FUN, ...)
```
The function `FUN` is called once for each element of `X`, with the
element as the first argument. The `...` is for additional arguments
to `FUN`, which are held constant across all calls.


Examples:
```{r}

```

`lapply()` always returns the result as a list.

"l" for **list** result.



The `sapply()` Function
---------------------

`sapply()` simplifies the result to a vector, when possible.

"s" for **simplified** result.

Examples:
```{r}

```

The `sapply()` function is useful if you are working interactively.


The `vapply()` Function
---------------------

`vapply()` simplifies the result to a vector of a specific data type.

"v" for **vector** result.


Examples:
```{r}

```

The `vapply()` function is more robust than `sapply()` because the
return type is guaranteed. Use `vapply()` when you write functions or
other non-interactive code.


The Split-Apply Pattern
-----------------------

The `split()` function splits a vector or data frame into groups
based on some other vector.

Split weight of dogs by the group column:
```{r}

```

The `split()` function is especially useful when combined with
`lapply()` or `sapply`().

```{r}

```
This is an R idiom!


The `tapply()` function is equivalent to the `split()` and `sapply()`
idiom.

"t" for **table**, because `tapply()` is a generalization of the
frequency-counting function `table()`.


Examples:
```{r}

```

The split-apply pattern is the base R equivalent to dplyr's
`group_by()` and `aggregate()` functions.

Like joins, this technique is important for analyzing tabular data
regardless of what programming language or packages you're using.


Choosing An Apply Function
--------------------------

1. Is the function you want to apply vectorized?

   If yes, use vectorization.

   Otherwise, continue to #2.


2. Do you want to apply the function to elements or to groups?

   For elements, continue to #3.

   For groups, use the split-apply pattern. Use `split()`, then
   continue to #3 to choose an apply function.

   Note `tapply()` is equivalent to `split()` and `sapply()`.


3. Will the function return the same data type for each element?

   If yes, continue to #4.

   Otherwise, use `lapply()`.


4. Are you working interactively?

   If yes, use `sapply()`.

   Otherwise, use `vapply()`.


Other Apply Functions
---------------------

* `apply()` applies a function to rows, columns, or elements of a
  matrix, and returns a matrix.
* `eapply()` applies a function to the elements of an environment.
* `mapply()` applies a function to each element of multiple vectors
  or lists.
* `rapply()` applies a function to all elements of a recursive list
  (a list of lists).

See this StackOverflow Post for a good summary:

    https://stackoverflow.com/a/7141669


