---
title: "Stat 33A - Lecture 9"
date: April 13, 2020
output: pdf_document
---


Review: Vectorization
=====================

A "vectorized" function is one that automatically applies an
operation to each element of a vector.

We saw this earlier with arithmetic:
```{r}
c(1, 2) + c(-1, 4)

sin(c(0, 1, 2))

#c(sin(0), sin(1), sin(2))

# Not vectorized:
mean(c(1, 2, 3))

mydata = list(1:3, c(0.1, -0.1, NA, 0.2))
mean(mydata[[1]])
mean(mydata[[2]], na.rm = TRUE)
```

Important takeaway: vectorized functions provide a way to do the same
operation repeatedly.



Apply Functions
===============

Doing the same operation repeatedly is a common pattern in
programming.

Vectorization is one way, but not all functions are vectorized.

In R, the "apply functions" are another way to do something
repeatedly.

The apply functions call a function on each element of a vector or
list.


The `lapply()` Function
---------------------

The first and most important apply function is `lapply()`. The syntax
is:
```
lapply(X, FUN, ...)
```
The function `FUN` is called once for each element of `X`, with the
element as the first argument. The `...` is for additional arguments
to `FUN`, which are held constant across all calls.


Examples:
```{r}
sin(c(0, 1, 2)) # do this

lapply(c(0, 1, 2), sin) # not this

# But for functions that aren't vectorized:
mydata
lapply(mydata, mean, na.rm = TRUE)

dogs = readRDS("data/dogs/dogs_full.rds")
dogs
lapply(dogs, class)
class(dogs$breed)
class(dogs$group)

```

`lapply()` always returns the result as a list.

"l" for **list** result.



The `sapply()` Function
---------------------

`sapply()` simplifies the result to a vector, when possible.

"s" for **simplified** result.

Examples:
```{r}
mean(c(1, 2))

lapply(mydata, mean, na.rm = TRUE)
sapply(mydata, mean, na.rm = TRUE)
class(sapply(mydata, mean, na.trm = TRUE))


sapply(mydata, mean, na.rm = TRUE)
sapply(mydata, median, na.rm = TRUE)

mean_median = function(x) {
   c(mean = mean(x, na.rm = TRUE), median = median(x, na.rm = TRUE))
}
sapply(mydata, mean_median)

sapply(mydata, function(x) {
   c(mean = mean(x, na.rm = TRUE), median = median(x, na.rm = TRUE))
})


sapply(dogs, summary)

```

The `sapply()` function is useful if you are working interactively.


The `vapply()` Function
---------------------

`vapply()` simplifies the result to a vector of a specific data type.

"v" for **vector** result.


Examples:
```{r}
sapply(mydata, mean, na.rm = TRUE)

vapply(mydata, mean, 0, na.rm = TRUE)

vapply(mydata, mean, "", na.rm = TRUE)

numeric(10)

vapply(mydata, mean, numeric(1), na.rm = TRUE)

quantile(mydata[[1]])

vapply(mydata, quantile, numeric(5), na.rm = TRUE)

integer(10)
character(10)
logical(10)
```

The `vapply()` function is more robust than `sapply()` because the
return type is guaranteed. Use `vapply()` when you write functions or
other non-interactive code.


The Split-Apply Pattern
-----------------------

The `split()` function splits a vector or data frame into groups
based on some other vector.

```{r}
x = c(1, 3, -7, 4, -2)
color = c("red", "blue", "red", "green", "blue")

split(x, color)
```


Split weight of dogs by the group column:
```{r}
split(dogs$weight, dogs$group)
```

The `split()` function is especially useful when combined with
`lapply()` or `sapply`().

```{r}
by_group = split(dogs$weight, dogs$group) # like group_by()
lapply(by_group, median, na.rm = TRUE)    # like summarize()

sapply(by_group, median, na.rm = TRUE)

sapply(by_group, quantile, na.rm = TRUE)
```
This is an R idiom!


The `tapply()` function is equivalent to the `split()` and `sapply()`
idiom.

"t" for **table**, because `tapply()` is a generalization of the
frequency-counting function `table()`.


Examples:
```{r}
tapply(dogs$weight, dogs$group, median, na.rm = TRUE)

tapply(dogs$weight, dogs$group, quantile, na.rm = TRUE)


table(dogs$group)

# The same as table():
tapply(dogs$group, dogs$group, length)
```

The split-apply pattern is the base R equivalent to dplyr's
`group_by()` and `summarize()` functions.

Like joins, this technique is important for analyzing tabular data
regardless of what programming language or packages you're using.


Choosing An Apply Function
--------------------------



1. Is the function you want to apply vectorized?

   If yes, use vectorization.

   Otherwise, continue to #2.


2. Do you want to apply the function to elements or to groups?

   For elements, continue to #3.

   For groups, use the split-apply pattern. Use `split()`, then
   continue to #3 to choose an apply function.

   Note `tapply()` is equivalent to `split()` and `sapply()`.


3. Will the function return the same data type for each element?

   If yes, continue to #4.

   Otherwise, use `lapply()`.


4. Are you working interactively?

   If yes, use `sapply()`.

   Otherwise, use `vapply()`.


Other Apply Functions
---------------------

* `apply()` applies a function to rows, columns, or elements of a
  matrix, and returns a matrix.
* `mapply()` applies a function to each element of multiple vectors
  or lists.
* `eapply()` applies a function to the elements of an environment.
* `rapply()` applies a function to all elements of a recursive list
  (a list of lists).

See this StackOverflow Post for a good summary:

    https://stackoverflow.com/a/7141669
