---
title: "Stat 33B - Lecture 12"
date: April 15, 2020
output: pdf_document
---

See the lecture slides for more notes.


The S3 Object System
====================

The `class()` function returns an object's S3 classes.
```{r}

```


Use `inherits()` to test whether an object is an instance of a
specific S3 class:
```{r}

```

Use `methods()` to list the methods defined for a given generic
function.

For S3 methods only, use `.S3methods()` instead.

```{r}

```

S3 methods are always named `GENERIC.CLASS()`.

Example: `print.lm()` is the `print()` method for the `lm` class.

But some functions with `.` in their name are not methods!

Example: `t.test()` is a generic function, not a method!

```{r}

```

Methods in packages are not always accessible from the global
workspace.

Use `getS3method()` or `getAnywhere()` to get these methods.

```{r}

```

Creating S3 Classes
-------------------

An S3 object is just an object with a "class" attribute.

So you can use `attr()`, `class()`, or `structure()` to create an S3
object:
```{r}

```

It's a good habit to define a constructor function:
```{r}

```
For S3, the constructor function implicitly defines the class.


Inheritance is based on the order of the classes:
```{r}

```


Creating S3 Generics & Methods
------------------------------

Use the `UseMethod()` function to create an S3 generic function.

S3 generics select a method based on one argument.

You can specify the argument to use in `UseMethod()`, or take the
default (the first argument).

Examples:
```{r}

```


Use the `NextMethod()` function to make a method call the method for
the object's parent class:
```{r}

```



The S4 Object System
====================

_Note: The S4 object system may appear on a homework or lab, but will not be on
any quizzes or exams._

Use `isS4()` to test whether an object is an S4 object:
```{r}

```

Use `is()` to test whether an object is an instance of a specific S3
or S4 class:
```{r}

```

Use `.S4methods()` to list the S4 methods defined for a given generic
function:
```{r}

```

Use `getClasses()` to list all defined S4 classes:
```{r}

```

Use `class?CLASSNAME` to get help with a specific class:
```{r}

```

Creating S4 Classes
-------------------

Use `setClass()` to create a new class:
```{r}

```

Use `new()` to create an object of a given class:
```{r}

```

S4 classes can inherit from built-in data types such as vectors and
lists.

The `.Data` slot provides access to the inherited object:
```{r}

```



Creating S4 Generics & Methods
------------------------------

Use `setGeneric()` to mark a function as an S4 generic:
```{r}

```

If the generic isn't already defined, you must provide a definition.

The standard definition is a call to `standardGeneric()`:
```{r}

```
This is the S4 equivalent of S3's `UseMethod()`.


Use `setMethod()` to define a method:
```{r}

```

Use the `callNextMethod()` function to call the method for the
object's parent class:
```{r}

```
