---
title: "Stat 33B - Lecture 12"
date: April 15, 2020
output: pdf_document
---

See the lecture slides for more notes.


The S3 Object System
====================

The `class()` function returns an object's S3 classes.
```{r}
class(5)

attr(5, "class")
```


Use `inherits()` to test whether an object is an instance of a
specific S3 class:
```{r}
inherits(5, "numeric")
```

Use `methods()` to list the methods defined for a given generic
function.

For S3 methods only, use `.S3methods()` instead.

```{r}
methods(print)

.S3methods(print)

.S3methods(summary)

.S3methods(t.test)
```

S3 methods are always named `GENERIC.CLASS()`.

Example: `print.lm()` is the `print()` method for the `lm` class.

But some functions with `.` in their name are not methods!

Example: `t.test()` is a generic function, not a method!

```{r}
print.lm
```

Methods in packages are not always accessible from the global
workspace.

Use `getS3method()` or `getAnywhere()` to get these methods.

```{r}
getS3method("print.lm")

getAnywhere("print.lm")
```

Creating S3 Classes
-------------------

An S3 object is just an object with a "class" attribute.

So you can use `attr()`, `class()`, or `structure()` to create an S3
object:
```{r}
x = 5
attr(x, "class") = "Foo"

x

class(x)

inherits(x, "Foo")
```

It's a good habit to define a constructor function:
```{r}
Animal = function(name = NA, breed = NA, weight = NA) {
  ani = list(name = name, breed = breed, weight = weight)
  structure(ani, class = "Animal")
}

Cat = function(...) {
  cat = Animal(...)
  class(cat) = c("Cat", class(cat))
  cat
}
```
For S3, the constructor function implicitly defines the class.


Inheritance is based on the order of the classes:
```{r}

```


Creating S3 Generics & Methods
------------------------------

Use the `UseMethod()` function to create an S3 generic function.

S3 generics select a method based on one argument.

You can specify the argument to use in `UseMethod()`, or take the
default (the first argument).

Examples:
```{r}
# Generic function:
move = function(animal) {
  UseMethod("move")
}

# "Animal" method of move():
move.Animal = function(animal) {
  message("The animal moved!")
}

move.Cat = function(animal) {
  message("The cat moved!")
}

move.default = function(animal) {
  message("The unknown class moved!")
}

move(f)

x = structure(5, class = "Foo")
move(x)
```


Use the `NextMethod()` function to make a method call the method for
the object's parent class:
```{r}
move.Cat = function(animal) {
  message("The cat moved!")
  NextMethod()
}

move(f)

methods(print)

print.Cat = function(x, ...) {
  msg = sprintf("%s is a %s that weights %i", x$name, x$breed, x$weight)
  message(msg)
}
```



The S4 Object System
====================

_Note: The S4 object system may appear on a homework or lab, but will not be on
any quizzes or exams._

Use `isS4()` to test whether an object is an S4 object:
```{r}
isS4(5)
```

Use `is()` to test whether an object is an instance of a specific S3
or S4 class:
```{r}
is(f, "Cat")
is(f, "Animal")
```

Use `.S4methods()` to list the S4 methods defined for a given generic
function:
```{r}
.S4methods("show")
```

Use `getClasses()` to list all defined S4 classes:
```{r}
getClasses()
```

Use `class?CLASSNAME` to get help with a specific class:
```{r}
class?ANY

class?environment

class?EmptyMethodsList

class?density # no documentation for "density" class
```


Creating S4 Classes
-------------------

Use `setClass()` to create a new class:
```{r}
setClass("Dog", slots = list(
  name = "character",
  breed = "character",
  weight = "numeric"
))
```

Use `new()` to create an object of a given class:
```{r}
sparky = new("Dog", name = "Sparky", breed = "Silky Terrier", weight = 12)

Dog = function(name = NA_character_, breed = NA_character_, weight = NA_real_) {
  new("Dog", name = name, breed = breed, weight = weight)
}

sparky@name
```

S4 classes can inherit from built-in data types such as vectors and
lists.

The `.Data` slot provides access to the inherited object:
```{r}
setClass("MyList", contains = "list")

x = new("MyList")
x@.Data

x@.Data[1] = "Hi"
x
```



Creating S4 Generics & Methods
------------------------------

Use `setGeneric()` to mark a function as an S4 generic:
```{r}
setGeneric("bark", function(dog) standardGeneric("bark"))
```

If the generic isn't already defined, you must provide a definition.

The standard definition is a call to `standardGeneric()`.

This is the S4 equivalent of S3's `UseMethod()`.


Use `setMethod()` to define a method:
```{r}
setMethod("bark", "Dog", function(dog) {
  message("Woof woof!")
})
```

Use the `callNextMethod()` function to call the method for the
object's parent class:
```{r}
# A better setup for the Animal/Dog/Cat objects:
setClass("Animal", slots = list(
  name = "character",
  breed = "character",
  weight = "numeric"
))

setClass("Dog", contains = "Animal")
setClass("Cat", contains = "Animal")

setGeneric("move", function(x) standardGeneric("move"))

setMethod("move", "Animal", function(x) {
  message("The animal moves.")
})

setMethod("move", "Cat", function(x) {
  callNextMethod()
  message("This is a cat, so it moves very, very quietly!")
})

mid = new("Cat", name = "Midnight", breed = "Norwegian Forest", weight = 8)
sparky = new("Dog", name = "Sparky", breed = "Silky Terrier", weight = 12)

move(mid)

move(sparky)
```
