---
title: "Stat 33A - Lecture 8"
date: April 6, 2020
output: pdf_document
---


The `%in%` Operator
===================

The `==` operator compares equality elementwise:
```{r}
x = c(1, 7, 9)
y = c(2, 7, 9)

x == y
```

If the vectors have different lengths, one will be recycled:
```{r}
x = c(1, 2, 1, 4)
y = c(1, 2)#1  2

x == y
```

If you want to check whether the elements on the left-hand side are
equal to ANY of the elements on the right-hand side, use `%in%`:
```{r}
x = c(1, 2, 1, 4)
y = c(2, 1)#2  1

x %in% y

x == y

# city %in% c("Oakland", "San Francisco")
# 
# city == "Oakland" | city == "San Francisco"
```


Writing Functions
=================

A function is like a factory: raw materials (inputs) go in, products
(outputs) come out.

```
         +-------+
-- in -->|   f   |-- out -->
         +-------+
```

You can write your own function with the `function` command.

The inputs to the function are called "parameters".

The output is called the "return value".

Just like any other object in R, you can assign functions to
variables.

```{r}
square = function(x) x^2 #+ 10

square(4)

square(c(10, 3))

is_negative = function(x)
   x < 0

is_negative(-1)

is_negative(12)
```

The code inside the function is called the "body" of the function.

Always indent the body by 2 spaces, so your code is easy to read!



Use curly braces `{ }` to write a function that contains more than
one line of code.

The result of the last line is automatically returned:
```{r}
sqrt_line_axb = function(x, a, b) {
   result = a * x + b
   sqrt(result)
}

sqrt_line_axb(0, 100, 4)

sqrt_line_axb(1, 100, 44)
```

Use `return()` to return a value early and IMMEDIATELY EXIT the
function:
```{r}
f = function() {
   x = sqrt(4 + 6)
   return(x)
   x + 10
}

f()
sqrt(10)
```
More on why this is useful near the end of the lecture.

To return multiple values, use a vector or list:
```{r}
f = function(x, y) {
   c(x + y, x * y)
}

f(5, 6)
```



Arguments in a call get assigned to parameters.

The order of the arguments matters:
```{r}
mean(c(1, 2, 3, NA), na.rm = TRUE)

tolerance_mean = function(x, tol) {
   mean(x[x > tol])
}

tolerance_mean(tol = 0.1, x = c(1, 2, 0.01))
```

Arguments can also be assigned to parameters by name.

It's possible to set "default arguments" for parameters.

Then the parameter is optional:
```{r}
tolerance_mean = function(x, tol = 0.1) {
   mean(x[x > tol])
}

tolerance_mean(c(1, 2, 0.015), tol = 0.01)
```

How to Use Functions
--------------------

Functions are the building blocks for solving problems.

Take a divide-and-conquer approach, breaking large problems into
smaller steps.

Use a short function for each step. Then it's easier to:

* Test that each step works correctly.
* Modify, reuse, or repurpose a step.

Before you write a function, it's useful to go through several steps:

1. Write down what you want to do, in detail. It can also help to
   draw a picture of what needs to happen.

2. Check whether there's already a built-in function. Google it!

3. Write the code to handle a simple case first. For data science
   problems, use a small dataset at this step.



How R Finds Variables
---------------------

The "scope" of a variable is the section of a program where that
variable exists and can be accessed.

Variables defined inside of a function are "local" to that function.

They cannot be accessed from outside the function:
```{r}
f = function() {
   z = 4
}

f()
z
```

Local variables reset each time a function is called:
```{r}
f = function() {
   message(z)
   z = 4
}

f()
```

A function can use variables defined outside (non-local), but only if
those variables are in scope where the function was **defined**.

This property is called "lexical scoping":
```{r}
x = 20
f = function(y) {
   y
}

f(x)
```

Variables defined directly in the R console are "global" and
available to any function.


Local variables "mask" or hide non-local variables with the same
name:
```{r}
x = 20
g = function() {
   x = 4
   x
}

g()
```

Locals get priority!

EXCEPTION: R ignores non-functions when looking up the name of a
called function.

For example:
```{r}

h = function(x) {
   mean = 0
   mean(x)
}

h(1:3)
```


In summary:

* A function's local variables are private.

* A function's local variables get reset for each call.

* A function can read non-local variables (in the scope where it was
  defined).

* Local variables mask non-local variables. Exception: calling
  non-local functions.


One final rule!

Variables are only looked up when they are actually used.

For a function, this means variables are only looked up when the
function is called.

This is called "dynamic lookup".

For example:
```{r}
f = function() z

z = 4
f()

z = 5
f()
```


Conditionals
============

Sometimes we need code to do different things depending on a
condition.

If-statements provide a way to write conditional code.

For example, suppose we want to print "Hello" if `x` is greater than
5, and print "Goodbye" otherwise:
```{r}
x = 5

if (x > 5) {
   message("Hello")
}

# else if (x == 5) {
#    message("x is 5")
# } else {
#    message("Goodbye")
# }
```

As with functions, indent the body of the if-statement.


Curly braces `{ }` are not strictly required in if-statements.

But you have to be careful if you don't use them:
```{r}
if (x > 5) message("Hello") else message("Goodbye")
```

An if-statement returns the final value in each branch:
```{r}
x = 10
if (x == 5) 3 else x
```

You can use this to do conditional assignment:
```{r}
y = if (x == 5) 3 else x

if (x == 5) {
   y = 3
} else {
   y = x
}
```

The condition in the if-statement should be a scalar:
```{r}
x = c(1, 2, 3)

if (x > 2) {
   message("Hello")
} else {
   message("Goodbye")
}
```


Parallel Vectors
----------------

Suppose we want to transform a vector `x` so that:

* Negative elements are set to 0.
* Positive elements are squared.

This is hard to do with if-statements because they take a scalar
condition.
```{r}
x = c(-1, -3, 10, 4, -2)

is_neg = x < 0

# NO GOOD:
#if (x < 0) {
#   
#}
```

Instead, we can use a "parallel vectors" strategy.

The idea is to work with two (or more) vectors:

1. An input vector of values to transform.

2. A group vector that identifies the different groups to transform
   within the input vector.

3. Optionally, an output or result vector to store the results.

The vectors are "parallel" because they must have the same length and
the elements must correspond.

In the example:
```{r}
x[is_neg] = 0
x[!is_neg] = x[!is_neg]^2
```


As another example, suppose we have two vectors:
```{r}
y = c(1, 2, 3, 4)
colors = c("red", "blue", "green", "blue")

# For y:
# Square red values
# Square root of blues
# Add 1 to greens

y[colors == "red"] = y[colors == "red"]^2
y[colors == "blue"] = sqrt(y[colors == "blue"])
y[colors == "green"] = y[colors == "green"] + 1
```


The `ifelse()` Function
-----------------------

R also has a vectorized `ifelse()` function.

For example:
```{r}
x = c(-1, -2, 3, 4)
ifelse(x < 0, x^2, x)
```

The `ifelse()` function is less efficient than a regular if-statement
or the parallel vectors strategy!
