---
title: "Stat 33B - Lecture 13"
date: April 22, 2020
output: pdf_document
---

Profiling
=========

The `Rprof()` function enables or disables R's built-in profiler.

Key parameters:

* `filename` -- path to output file. Defaults to `Rprof.out`.
* `interval` -- time between samples, in seconds.
* `memory.profiling` -- whether to track memory in addition to time.

Call `Rprof(NULL)` to disable the profiler.



The `summaryRprof()` function summarizes results from `Rprof()`.

Key parameters:

* `filename` -- path to output file. Defaults to `Rprof.out`.
* `memory` -- how to display memory profile. Use `"both"` to see
  total changes.

The summary lists times in seconds and memory in bytes.


The profvis package provides a visualization, making it easier to
understand the results.

Call `profvis()` on the code you want to profile.

Under the hood, profvis still uses `Rprof()`.



Example: Rejection Sampler
--------------------------

Recall the rejection sampler from homework 4.

Preallocation was important here.

What does the profiler show if we don't preallocate?

```{r}
library(profvis)

profvis({
dslide = function(x) {
  ifelse(x > 1, 0,
    ifelse(x > 0, dnorm(x) / dnorm(0), dunif(x, -1, 0))
  )
}

# PREALLOCATION:
rslide = function(n) {
  samp = numeric(n)
  accepted = 0
  while (accepted < n) {
    x = runif(1, -1, 1)
    y = runif(1)
  
    if (y < dslide(x)) {
      accepted = accepted + 1
      samp[accepted] = x
    }
  }

  samp
}

rslide(1e5)
})
```

```{r}
summaryRprof()
```

Example: Random Walk
--------------------

In a 2-dimensional random walk, at each step we randomly choose a
move in one of the four compass directions.

Goal: write an efficient function to compute the path of a 2-d random
walk.

```{r}
profvis({
  
rwalk = function(n) {
  x = numeric(n + 1)
  y = numeric(n + 1)
  
  axis = sample(c(0, 1), n, replace = TRUE) == 0
  move = sample(c(-1, 1), n, replace = TRUE)
  
  # for i in 2 to n+1
  for (i in seq_len(n) + 1) {
    if (axis[i - 1]) { # x
      x[i] = x[i - 1] + move[i - 1]
      y[i] = y[i - 1]
    } else { # y
      x[i] = x[i - 1]
      y[i] = y[i - 1] + move[i - 1]
    }
  }
  
  data.frame(x = x, y = y)
}

rwalk(1e6)
})
```

```{r}
profvis({
rwalk2 = function(n) {
  xmove = c(-1, 1, 0, 0)
  ymove = c(0, 0, -1, 1)
  move = sample(1:4, n, replace = TRUE)  
  
  x = c(0, cumsum(xmove[move]))
  y = c(0, cumsum(ymove[move]))
  
  data.frame(x = x, y = y)
}

rwalk2(1e6)
})
```
