---
title: "Stat 33B - Lecture 14"
date: April 29, 2020
output: pdf_document
---

Review: Creating Closures
=========================

How do you create a function that "remembers" something between
calls?

Same skeleton every time:
```{r}

```

Example:
```{r}

```

Review: Creating S3 Objects
===========================

How do you create an S3 object?

Most of the time, use a list as the base for the object:
```{r}

```

Add named elements to the list to create fields:
```{r}

```

Formulas
========

Examples:
```{r}

```


dplyr
=====

Remember the dogs data?
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")

head(dogs)
```


Use `slice()` to choose rows by position:
```{r}

```

Use `filter()` to choose rows that satisfy a condition:
```{r}

```

Suppose we want all toy dogs and all sporting dogs:
```{r}

```

Suppose we want to get all dogs with group "toy" and weight < 10:
```{r}

```


Use `select()` to choose columns by name or position:
```{r}

```

Use `:` to indicate a range of rows or columns, including by name:
```{r}

```

Use `-` to exclude rows or columns, including by name:
```{r}

```

Other useful `dplyr` functions:

* `arrange()` changes the ordering of the rows.
* `mutate()` adds new columns by transforming existing columns.
* `group_by()` splits rows into groups when summarizing.
* `summarise()` reduces multiple rows down to a single value.


magrittr
========

The pipe operator is different way to call a function:
```{r}

```

Use `.` as a placeholder for the previous operand:
```{r}

```


Non-standard Evaluation
=======================

The `substitute()` function is the key to NSE.

When called on a parameter inside of a function, `substitute()`
returns the unevaluated argument to that parameter:
```{r}

```

These objects that represent code are called "language objects".

They have a tree structure.

Use the subset `[` operator to access or edit language objects:
```{r}

```

The `quote()` function captures its argument as a language object:
```{r}

```

The `eval()` function evaluates a language object:
```{r}

```

Non-standard evaluation is extremely powerful.

With great power comes great responsibility!

Think twice before creating a function that uses NSE.
