---
title: "Stat 33B - Lecture 14"
date: April 29, 2020
output: pdf_document
---

Review: Creating Closures
=========================

How do you create a function that "remembers" something between
calls?

Same skeleton every time:
```{r}
make_foo = function() {
  # Put variables that will persist between calls here:
  memory = numeric(0)
  
  # This is the foo() function:
  function(x) {
    # Code to do computions for foo()
    # ...
    # memory[i] <<- x 
  }
}

foo = make_foo()

# Then:
foo(3)
foo(4)
```

Example:
```{r}
# Create a function that returns the last value passed as an argument!
make_remember = function() {
  memory = NA
  
  function(x) {
    old_mem = memory
    memory <<- x 
    old_mem
  }
}

remember = make_remember()


remember("Hi") # NA
remember("Hello") # "Hi"
```

Review: Creating S3 Objects
===========================

How do you create an S3 object?

Most of the time, use a list as the base for the object:
```{r}
structure(list(), class = "foo")
```

Add named elements to the list to create fields:
```{r}
foo = function() {
  base = list(x = 4)
  structure(base, class = "foo")
}

obj = foo()
obj$x
```

Formulas
========

Remember the dogs data?
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")

head(dogs)
```


Examples:
```{r}
plot(datadog ~ popularity, data = dogs)

lm(datadog ~ popularity + weight, data = dogs)
```


dplyr
=====

Use `slice()` to choose rows by position:
```{r}
library(dplyr)

slice(dogs, 1:3)
# dogs[1:3, ]
```

Use `filter()` to choose rows that satisfy a condition:
```{r}
filter(dogs, group == "herding")
# dogs[dogs$group == "herding", ]
# subset(dogs, group == "herding")
```

Suppose we want all toy dogs and all sporting dogs:
```{r}
filter(dogs, group %in% c("toy", "sporting"))
```

Suppose we want to get all dogs with group "toy" and weight < 10:
```{r}
filter(dogs, group == "toy" & weight < 10)
filter(dogs, group == "toy", weight < 10)
```


Use `select()` to choose columns by name or position:
```{r}
select(dogs, group)
#dogs["group"]

select(dogs, 3)
```

Use `:` to indicate a range of rows or columns, including by name:
```{r}
select(dogs, 2:4)

# ONLY IN dplyr:
select(dogs, breed:popularity)
```

Use `-` to exclude rows or columns, including by name:
```{r}
select(dogs, -1)
dogs[-1]

# ONLY IN dplyr:
select(dogs, -c(breed, datadog, popularity))
```

Other useful `dplyr` functions:

* `arrange()` changes the ordering of the rows.
* `mutate()` adds new columns by transforming existing columns.
* `group_by()` splits rows into groups when summarizing.
* `summarise()` reduces multiple rows down to a single value.


magrittr
========

The pipe operator is different way to call a function:
```{r}
c(1, 2, 3) %>% sum %>% sqrt

sqrt(sum(c(1, 2, 3)))

total = sum(c(1, 2, 3))
sqrt(total)
```

Use `.` as a placeholder for the previous operand:
```{r}
TRUE %>% mean(c(1, 2, 3, NA), na.rm = .)
```


Non-standard Evaluation
=======================

Example of `curve()`:
```{r}
curve(sqrt(x - 1), 0, 2)
```


The `substitute()` function is the key to NSE.

When called on a parameter inside of a function, `substitute()`
returns the unevaluated argument to that parameter:
```{r}
f = function(x) {
  substitute(x)
}

z = f(4 + 10)
z

z = f(g(10) + 3)
```

These objects that represent code are called "language objects".

They have a tree structure.

Use the subset `[` operator to access or edit language objects:
```{r}
z[[2]]

z[[2]][[2]] = 10

class(z)
sapply(as.list(z), class)
as.list(z)
```

The `quote()` function captures its argument as a language object:
```{r}
x = quote(5 + 6)
```

The `eval()` function evaluates a language object:
```{r}
eval(x)
eval(z)
```

Non-standard evaluation is extremely powerful.

With great power comes great responsibility!

Think twice before creating a function that uses NSE.
