---
title: "Stat 33B - Lecture 3"
date: February 5, 2020
output: pdf_document
---

Announcements
-------------

Second homework assignment will be posted to the bCourse later today.

From now on:

* The attendance question will be at the beginning of lecture.
* If you're late, ask me for it at the end of lecture.
* The attendance question will no longer be posted in the notes.


Review
------

Last time we learned about types and classes.
```{r}
typeof(5)

class(5)
```

We also saw data frames, lists, matrices, and arrays.
```{r}
list(4, "hi")

matrix(1:6, 2, 3)
```
R stores matrices in column order.


# Data Types

Attributes
----------

Attributes are name-value pairs in the metadata of an R object.

Several functions for working with attributes:

* `attr()` to get or set one attribute on an object
* `attributes()` to get all attributes on an object
* `structure()` to set all attributes on an object

We've already been using attributes.
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
class(dogs)
```


Classes are stored in the "class" attribute:
```{r}
attr(dogs, "class")

attributes(dogs)

attr(dogs, "names")
```

The `class` function is just a helper for this attribute.

Names are also attributes:
```{r}
names(dogs)
```

Some common attributes:

* `class` -- class
* `row.names` -- row names
* `names` -- element names or column names
* `dim` -- dimensions (on matrices)
* `dimnames` -- names of dimensions (on matrices)

```{r}
x = matrix(1:4, 2, 2)
attributes(x)
```


Some of these also have helper functions:
```{r}
dim(x)
```


The `length` function is superficially similar to these, but there's no length
attribute.

```{r}
length(x)
```


Idiomatic R: use a helper function to set attributes when possible.


We can set the class of an object with the helper function:
```{r}
x = 5
class(x) = "data.frame"
x
```

Or with the `attr()` or `structure()` functions:
```{r}
attr(x, "class") = "data.frame"

structure(5, class = "data.frame")
```

Q: What happens if there's a type in the class name?

A: Not an error, but R will probably not recognize the class:
```{r}
x = structure(5, class = "dat.frame")
x
```
In a few lectures, we'll see how to create our own classes.


Notice that "atomic" values like numbers don't have attributes:
```{r}
attributes(5)
```

But the class function always returns a class:
```{r}
class(5)
```

The `attr()` function shows that the class for this object is not an attribute:
```{r}
attr(5, "class")
```
Instead, the class for atomic objects is built into R.


Special Values
--------------

R has four special values.

`NA` represents a missing or unknown value in a data set.
```{r}
class(NA)
typeof(NA)
```


The missing value `NA` is contagious!
```{r}
5 + NA
```

Using a unknown argument in a computation usually produces an unknown result.


`NULL` represents a value that's not defined _in R_. Functions often use `NULL`
to mean the absence of a result.
```{r}
attr(5, "class")
```

For instance, if we try to get the matrix dimensions of a vector:
```{r}
dim(6)
```

`NaN`, or "not a number", represents a value that's not defined mathematically.
Produced by some computations:

```{r}
0 / 0
```


`Inf` represents infinity. Produced by some computations:

```{r}
-5 / 0
```


Comparison Operators
--------------------

We saw operators for doing arithmetic: `+`, `-`, `*`, `/`, `^`

R also has operators for making comparisons. They are:

* `==` for equality
* `<`, `<=`, `>`, `>=` for inequality

Like the arithmetic operators, these are vectorized.

Examples:
```{r}
c(1, 2, 3) == c(2, 2, 2)

c(1, 2, 3) < c(2, 2, 2)

# NA is contagious
c(NA, 3) < c(2, 3)
```

To check equality for special values, use the `is.` functions:

* `is.na()`
* `is.null()`
* `is.nan()`
* `is.infinite()`

Examples:
```{r}
# NA is contagious
NA == NA

is.na(c(NA, 1))
```


To check equality within a tolerance, use `all.equal()`. Useful for numerical
algorithms.

```{r}
isTRUE(all.equal(4, 5, 6))
```


To check whether two objects are identical, use `identical()`. Useful for
non-vectorized equality checks.

```{r}
c(1, 2, 3) == c(1, 2, 3)

identical(c(1, 7, 3), c(1, 2, 3))
```


Logical Operators
-----------------

R also has operators for combining logical values. They are:

* `!` for negation ("NOT")
* `&` for conjunction ("AND"). `TRUE` when both arguments are `TRUE`.
* `|` for disjunction ("OR"). `TRUE` when at least one argument is `TRUE`.
* `xor()` for exclusive disjunction ("XOR"). `TRUE` when exactly one argument is
  `TRUE`.

These four operators are vectorized.

Examples:
```{r}
!c(TRUE, FALSE)

c(TRUE, TRUE) & c(TRUE, FALSE)
```

There are also short-circuited versions of `&` and `|`:

* `&&`
* `||`

These are __not__ vectorized!

Short-circuited means these operators only evaluate the second argument if it
is necessary to determine the result.

For instance, in `TRUE || x` we don't need to know the value of `x` to know the
result will be `TRUE`. So in this case, `||` never looks up the value of `x`.

Examples:
```{r}
# No need to evaluate the second FALSE:
FALSE && FALSE


# Easy source of bugs with vectors, since they are not vectorized:
c(FALSE, TRUE) && c(TRUE, TRUE)
```


The `any()` and `all()` functions check whether any or all elements of a
logical vector are `TRUE`, respectively:
```{r}
any(c(FALSE, FALSE, FALSE))


all(c(FALSE, TRUE))
```

It is idiomatic to use `sum()` to count `TRUE` values in a vector:
```{r}
sum(c(TRUE, FALSE, TRUE)) # faster

# Another way:
table(c(TRUE, FALSE, FALSE)) # slower, but more information
```


Factors
-------

We need the dogs data again:
```{r}
dogs = readRDS("data/dogs/dogs_full.rds")
```

R represents categorical data using the class `factor`:
```{r}
class(dogs$group)
```

Factors are not strings!

For efficiency, R stores factors as integer vectors:
```{r}
typeof(dogs$group)
```

Each integer corresponds to one category or "level" of the factor.

Besides efficiency, R uses factors to keep track of _possible_ categories even
if they are not actually present in the vector.

Functions for working with factors:

* `factor()` creates a factor.
* `levels()` returns the levels of a factor.
* `droplevels()` removes levels of a factor that are not present.
* `table()` prints a count of elements of each level. Also useful for other
  data types.

Examples:
```{r}
levels(dogs$group)

factor(c("hi", "bye", "hi"))

# Second argument sets the levels:
hair = factor(c("black", "blonde"), c("black", "blonde", "brown", "red"))

# Get counts for every level, even ones that weren't observed:
table(hair)
```


# Working with Data


Taking Subsets
--------------

Taking subsets is a fundamental operation for working with data in R!

Two primary operators for taking subsets in R:

1.  `[[` to "extract" an element.

    + Selects exactly one element.
    + Drops the container.
    + By position or name.

2. `[` to "subset" an element.

    + Selects any number of elements.
    + Keeps the container.
    + By position, name, or logical values.


Let's test these out on a vector:
```{r}
x = c(2, -1, 6)
```

Subsetting by position (positions start at 1 in R):
```{r}
x[1]

x[[1]]

x[c(2, 1)]

# Can't get multiple elements with [[
x[[c(2, 1)]]
```

Negative positions mean "everything except":
```{r}
x[-c(1, 3)]
```

-----

_This section wasn't covered in the lecture and will be covered next week
instead. I'm including it here for anyone that wants to read ahead._

Subsetting by name:
```{r}
y = c(a = 1, b = 2)

y["a"]

x[["a"]]

y[c("a", "a")]
```


Subsetting by logical value:
```{r}
x < 2 # logical vector


# Length of the logical vector must match.
#
# TRUE to keep an element, FALSE to drop an element.
x[x < 2]


# Invalid with [[
x[[x < 2]]
```


`[[` drops the container, `[` does not.

This is most apparent for lists:

```{r}
mylist = list(a = 8, b = 44)

mylist[1]
class(mylist[1])

mylist[[1]]
class(mylist[[1]])
```

A visual representation of this idea:
<https://twitter.com/hadleywickham/status/643381054758363136>


Why do `[` and `[[` seem to do the same thing for vectors?

Vectors contain scalars, but in R, a scalar is a 1-vector. So nothing to
"extract" with `[[`.


Data frames are lists, so:
```{r}
dogs[[1]]

dogs[1]
```

We can also do multi-dimensional subsetting:
```{r}
dogs[[1, 1]]

dogs[1, 1]
```

Use a blank to mean "everything" along one dimension:
```{r}
dogs[1, ]
```

The `[` has a parameter called `drop`:
```{r}
?`[`
```

The default `drop = TRUE`, means that `[` will "drop" (remove) the dimensions
from objects that can be represented as a vector.

This is a potential source of bugs:
```{r}
x = matrix(1:4, 2, 2)

# Matrix multiplication operator is %*%
#
# (2x1) * (1x2) = (2x2)
x[, 1] %*% x[1, ] # wrong answer

# Try without drop:
x[, 1, drop = FALSE] %*% x[1, ]
```
Even some of the R developers regret that the default is not `drop = FALSE`!
